---
title: "Replication code for 'The Effects of Starlink adoption on forest degradation in the Amazon'"
subtitle: "Anonymized replication package for peer review"
output: 
  html_document: 
    self_contained: true
    mathjax: null
    pandoc_args:
      - --mathml
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false

date: "`r format(Sys.Date(), '%Y-%m-%d')`"
---

<style type="text/css">
/* ===== Full-width, browser-adaptive layout ===== */
html body div.main-container,
.main-container,
.container-fluid,
.container {
  max-width: 100% !important;
  width: auto !important;
  margin-left: 2% !important;
  margin-right: 2% !important;
  padding-left: 0 !important;
  padding-right: 0 !important;
}

/* HTML (kable / data.frame) tables span the full content width. */
table.table, table.dataframe, .kable_wrapper table { width: 100% !important; }

/* Wide CONSOLE-printed output (result tables use R.options width = 1200) and code
   blocks scroll horizontally instead of being clipped at the page edge. */
pre, pre > code, .sourceCode {
  white-space: pre !important;
  overflow-x: auto !important;
  word-break: normal !important;
  max-width: 100% !important;
}

/* Figures never overflow the container. */
img { max-width: 100%; height: auto; }
</style>


```{r, include=FALSE}
# Initializing code
rm(list = ls(all.names = TRUE))
gc()

# ---- data locations ----------------------------------------------------------
DATA_DIR = "../data/"

# Municipal polygons for all maps.
mun_legal_amazon_geometry = paste0(DATA_DIR, "processed/mun_zone.RDS")

# Controls
# Model II controls.
model_controls = c(
  "year:dens",
  "year:lat",
  "year:lon",
  "year:dist_coast",
  "year:dist_brasilia",
  "year:r_way_road_pav",
  "year:r_way_rail_oper",
  "tmax",
  "prcp",
  "geosat",
  "spei12_dec",
  "spei12_dec:pre2022_fire_scar_intensity",
  "soyield_ful_mun_high:soy_price_lagged_level",
  "pasture_suit_mun:cattle_price_lagged_level"
)

# The coverage-year instruments omit the 2022 reference year.
calendar_iv_names = paste0("ca_", 2023:2024)

# Parallel workers for the leave-one-out and leave-one-state-out sweeps.
robust_max_workers = min(8, parallel::detectCores() - 1)
```


```{r Initialization, include=FALSE}
# Loading required R packages
library(fixest)
fixest::setFixest_nthreads(1) # Use one fixest thread to avoid Windows OpenMP access violations.

library(tidyverse)
library(ggpubr)
library(future.apply)
```


```{r VariableNames, echo=F}
# Defining variables and names
var_names = rbind(
  # Degradation flags
  c("r_n_mun_deter_degr", "Degradation (flags per 100 km²)", "out"),
  c("r_n_mun_deter_degr_fire", "    fire scars", "out"),
  c("r_n_mun_deter_degr_wood", "    selective-cutting", "out"),
  c("r_n_mun_deter_degr_outr", "    other means", "out"),

  c("r_n_ti_deter_degr", "  in indigenous lands", "out"),
  c("r_n_uc_deter_degr", "  in conservation units", "out"),
  c("r_n_sett_deter_degr", "  in settlements", "out"),
  c("r_n_ndpf_deter_degr", "  in undesignated forests", "out"),
  c("r_n_sicar21_large_deter_degr", "  in large properties", "out"),
  c("r_n_over_ndpf_large_deter_degr", "    ∩ undesignated forest", "out"),
  c("r_n_sicar21_small_deter_degr", "  in small properties", "out"),
  c("r_n_over_ndpf_small_deter_degr", "    ∩ undesignated forest", "out"),


  # Degradation area
  c("r_area_mun_deter_degr", "Degradation (% forest area)", "out"),
  c("r_area_mun_deter_degr_fire", "    fire scars", "out"),
  c("r_area_mun_deter_degr_wood", "    selective-cutting", "out"),
  c("r_area_mun_deter_degr_outr", "    other means", "out"),

  c("r_area_ti_deter_degr", "  in indigenous lands", "out"),
  c("r_area_uc_deter_degr", "  in conservation units", "out"),
  c("r_area_sett_deter_degr", "  in settlements", "out"),
  c("r_area_ndpf_deter_degr", "  in undesignated forests",  "out"),
  c("r_area_sicar21_large_deter_degr", "  in large properties", "out"),
  c("r_area_over_ndpf_large_deter_degr", "    ∩ undesignated forest", "out"),
  c("r_area_sicar21_small_deter_degr", "  in small properties", "out"),
  c("r_area_over_ndpf_small_deter_degr", "    ∩ undesignated forest", "out"),

  # Fire ignitions
  c("rm_n_igni", "Fire ignitions (per 100 km² munic. area)", "out"),
  c("log_n_igni", "  log(count + 1)", "out"),
  c("ihs_n_igni", "  IHS of count", "out"),
  c("conc_igni_hhiunb", "  Ignition timing HHI", "out"),
  c("disp_shhiunb", "  Ignition spatial HHI", "out"),

  # Land cover change over the FIXED 2021 forest stock
  c("sh21_for2non", "Forest replaced (% 2021 forest)", "out"),
  c("sh21_veg", "  by natural non-forest area", "out"),
  c("sh21_agr_agr", "  by cropland", "out"),
  c("sh21_agr_pec", "  by pasture", "out"),
  c("sh21_agr_out", "  by mosaic/forest plantation", "out"),
  c("sh21_nvg_urb", "  by urban infrastructure", "out"),
  c("sh21_nvg_min", "  by mining", "out"),

  # Fines
  c("r_n_mun_police", "Enforcement fines (per 100 km²)", "out"),
  c("r_n_mun_police_faun", "    for fauna damage", "out"),
  c("r_n_mun_police_polu", "    for pollution", "out"),
  c("r_n_mun_police_admi", "    for admin. issues", "out"),
  c("r_n_mun_police_flor", "    for flora damage", "out"),
  c("r_n_mun_police_fire", "      use of fire", "out"),

  # Particulate matter
  c("pm10mun", "Coarse particulate matter (PM10)", "out"),
  c("pm25mun", "Fine particulate matter (PM2.5)", "out"),
  c("pm1mun", "Submicron particulate matter (PM1)", "out"),

  # Mortality
  c("mort_total", "Deaths per 100,000 inhab.", "out"),
  c("mort_extcause", "  external cause", "out"),
  c("mort_infect", "  infection", "out"),
  c("mort_neoplasm", "  neoplasm", "out"),
  c("mort_blood", "  blood disorders", "out"),
  c("mort_endocrine", "  endocrine/metabolic issues", "out"),
  c("mort_mental", "  mental disorders", "out"),
  c("mort_nervous", "  nervous system", "out"),
  c("mort_eyes", "  eye conditions", "out"),
  c("mort_ears", "  ear conditions", "out"),
  c("mort_circulatory", "  circulatory issues", "out"),
  c("mort_respiratory", "  respiratory issues", "out"),
  c("mort_digestive", "  digestive issues", "out"),
  c("mort_skin", "  skin conditions", "out"),
  c("mort_musculoskeletal", "  musculoskeletal issues", "out"),
  c("mort_genitourinary", "  genitourinary issues", "out"),
  c("mort_pregnancy", "  pregnancy/childbirth", "out"),
  c("mort_perinatal", "  perinatal causes", "out"),
  c("mort_malformation", "  malformations", "out"),
  c("mort_other", "  other diseases", "out"),


  # Deforestation (PRODES)
  c("r_area_mun_prodes_defo", "PRODES Deforestation (% forest area)", "out"),
  c("r_area_mun_prodes_defo_ccut", "    clear-cut (total)", "out"),
  c("r_area_mun_prodes_defo_cces", "    clear-cut with exposed soil", "out"),
  c("r_area_mun_prodes_defo_ccwv", "    clear-cut with vegetation", "out"),
  c("r_area_mun_prodes_defo_dpdp", "    by progressive degradation", "out"),
  c("r_area_mun_prodes_defo_mine", "    with mining patterns", "out"),

  # Deforestation (PRODES) - Flags
  c("r_n_mun_prodes_defo", "PRODES Deforestation (flags per 100 km²)", "out"),
  c("r_n_mun_prodes_defo_ccut", "    clear-cut (total)", "out"),
  c("r_n_mun_prodes_defo_cces", "    clear-cut with exposed soil", "out"),
  c("r_n_mun_prodes_defo_ccwv", "    clear-cut with vegetation", "out"),
  c("r_n_mun_prodes_defo_dpdp", "    by progressive degradation", "out"),
  c("r_n_mun_prodes_defo_mine", "    with mining patterns", "out"),


  # Covariates
  c("coverarea", "Mobile coverage (% munic. area)", "cov"),
  c("starlink", "Starlink per 1,000 inhab.", "cov"),
  c("geosat", "GEO internet per 1,000 inhab.", "cov"),
  c("spei12_dec", "SPEI-12 (December)", "cov"),
  c("spei12_jul", "SPEI-12 (July)", "cov"),
  c("pre2022_fire_scar_intensity", "Accumulated fire-scar area (% 2021 forest)", "cov"),
  c("spei12_dec:pre2022_fire_scar_intensity",
    "December SPEI-12 $\\times$ pre-2022 fire-scar intensity", "cov"),
  c("spei12_jul:pre2022_fire_scar_intensity",
    "July SPEI-12 $\\times$ pre-2022 fire-scar intensity", "cov"),
  c("areamun", "Municipal area (1,000 km²)", "cov"),
  c("area_forest", "   Forested area (1,000 km²)", "cov"),

  c("lat", "Latitude", "cov"),
  c("lon", "Longitude", "cov"),
  c("pop", "Population (1,000)", "cov"),
  c("dens", "Population density (pop./km²)", "cov"),
  c("dist_coast", "Distance to coast (1,000 km)", "cov"),
  c("dist_brasilia", "Distance to Brasilia (1,000 km)", "cov"),
  c("tmax", "Maximum temperature (°C)", "cov"),
  c("prcp", "Mean monthly precipitation (mm)", "cov"),

  # Agricultural rents
  c("soyield_ful_mun_high", "Soy yield potential (tons/ha)", "cov"),
  c("pasture_suit_mun", "Pasture suitability (0-1 index)", "cov"),
  c("soy_price_current", "Soybean price (BRL per 60 kg)", "cov"),
  c("cattle_price_current", "Cattle price (BRL per 15 kg)", "cov"),
  c("soy_price_lagged_level", "Lagged soybean price (BRL per 60 kg)", "cov"),
  c("cattle_price_lagged_level", "Lagged cattle price (BRL per 15 kg)", "cov"),

  # Paved-road and operating-railway densities
  c("r_way_road_pav", "Paved roads (km/100 km² munic.)", "cov"),
  c("r_way_rail_oper", "Operating railway density (km/100 km² munic.)", "cov"),

  # Heterogeneities
  c("uc_dummy", "conservation-unit presence", "het"),
  c("faction", "faction presence", "het")
) %>%
  as.data.frame()

colnames(var_names) = c("var", "varname", "type")

```

---

# Normal Year

**Calendar-year panel: Jan 2022 to Dec 2024.**

```{r DataNormalization, echo=F}
# Dataset and normalization
dta = readRDS(paste0(DATA_DIR, "dataset_normalyr.RDS")) %>%

  mutate() %>%
  filter(year >= 2017 & year <= 2024 & !is.na(coverarea)) %>%
  mutate(year = factor(year, levels = as.character(2024:2017))) %>%
  fastDummies::dummy_cols("year") %>%
  mutate(
    pasture_suit_mun = pasture_suit_mun / 100, # FGGD's 0-100 index is stored as 0 to 10000.
    way_rail_oper = way_rail,
    ca_2017 = year_2017 * coverarea,
    ca_2018 = year_2018 * coverarea,
    ca_2019 = year_2019 * coverarea,
    ca_2020 = year_2020 * coverarea,
    ca_2021 = year_2021 * coverarea,
    ca_2022 = year_2022 * coverarea,
    ca_2023 = year_2023 * coverarea,
    ca_2024 = year_2024 * coverarea,
    sh_for2non = 1 - sh_flo,

    # Cover transitions rebased on the FIXED 2021 forest stock
    rebase21 = lag_areaforest / area_forest,
    sh21_for2non = (1 - sh_flo) * rebase21,
    sh21_flo = sh_flo * rebase21,
    sh21_veg = sh_veg * rebase21,
    sh21_agr_pec = sh_agr_pec * rebase21,
    sh21_agr_agr = sh_agr_agr * rebase21,
    sh21_agr_out = sh_agr_out * rebase21,
    sh21_nvg_min = sh_nvg_min * rebase21,
    sh21_nvg_urb = sh_nvg_urb * rebase21,

    uc_dummy = ifelse(areauc > 0, 1, 0)
  ) %>%
  select(-starts_with("year_20"))


# Define tenure-presence samples
tenure_denom = c(
  ti = "areati",
  uc = "areauc",
  sett = "areasett",
  ndpf = "areandpf",
  sicar21_large = "areasicar21_large",
  sicar21_small = "areasicar21_small",
  over_ndpf_large = "areaover_ndpf_large",
  over_ndpf_small = "areaover_ndpf_small"
)

# Zero tenure areas → NA in tenure-level outcomes
for (tag in names(tenure_denom)) {
  denom_col = tenure_denom[[tag]]
  tenure_vars = colnames(dta)[grepl(
    paste0("^(area|n)_", tag, "_"),
    colnames(dta)
  )]
  dta = dta %>%
    mutate(across(
      all_of(tenure_vars),
      ~ if_else(dta[[denom_col]] == 0, NA_real_, .)
    ))
}

# Outcome normalization
# `r_` uses fixed 2021 forest area and `rm_` uses municipal area.
# Road density uses municipal area.

# Area outcomes
area_list = colnames(dta)[startsWith(colnames(dta), "area") | startsWith(colnames(dta), "way")]
for (y in area_list) {
  dta[[paste0("rm_", y)]] = (dta[[y]] / dta[["areamun"]]) * 100
  dta[[paste0("r_", y)]] = if (startsWith(y, "way")) {
    dta[[paste0("rm_", y)]]
  } else {
    (dta[[y]] / dta$area_forest) * 100
  }
}

# Count outcomes
n_list = colnames(dta)[startsWith(colnames(dta), "n_") | startsWith(colnames(dta), "offend_")]
for (y in n_list) {
  dta[[paste0("rm_", y)]] = (dta[[y]] / dta[["areamun"]]) * 100
  dta[[paste0("r_", y)]] = (dta[[y]] / dta$area_forest) * 100
}

# Ignition count transformations reported in the manuscript.
dta$log_n_igni = log(dta$n_igni + 1)
dta$ihs_n_igni = asinh(dta$n_igni)

# Forest transition outcomes
sh_list = colnames(dta)[grepl("^sh(m|21)?_", colnames(dta))]
for (y in sh_list) {
  dta[[y]] = dta[[y]] * 100
}

# Rescaling
dta = dta %>%
  mutate(
    areamun = areamun / 1000,
    area_forest = area_forest / 1000,
    dist_coast = dist_coast / 1000,
    dist_brasilia = dist_brasilia / 1000
  ) %>%
  # Degenerate normalizations: a denominator divided by itself.
  select(-any_of(c("r_areamun", "rm_areamun", "r_area_forest")))

dta_normal = dta
```

## Descriptive Statistics

```{r DescriptiveStatistics, R.options = list(width = 1200), echo=F}
# Descriptive statistics
vars = c(
  "r_n_mun_deter_degr",
  "r_n_mun_deter_degr_fire",
  "r_n_mun_deter_degr_wood",
  "r_n_mun_deter_degr_outr",
  
  "sh21_for2non", "sh21_veg", "sh21_agr_agr", "sh21_agr_pec", 
  "sh21_agr_out", "sh21_nvg_urb", "sh21_nvg_min",

  "rm_n_igni",

  "r_n_mun_police", "r_n_mun_police_admi", "r_n_mun_police_polu",
  "r_n_mun_police_faun", "r_n_mun_police_flor", "r_n_mun_police_fire",

  "pm10mun", "pm25mun", "pm1mun",

  "coverarea", "starlink", "geosat",
  "areamun", "area_forest", "lat", "lon",

  "pop", "dens", "dist_coast", "dist_brasilia",
  "r_way_road_pav", "r_way_rail_oper",
  "soyield_ful_mun_high", "soy_price_current",
  "pasture_suit_mun", "cattle_price_current",
  "tmax", "prcp", "spei12_dec"
)


# Format mean with standard deviation in parentheses.
fmt = function(x, digits = 2) {
  if (all(is.na(x))) {
    return(NA_character_)
  }
  m = mean(x, na.rm = TRUE)
  s = sd(x, na.rm = TRUE)
  sprintf(paste0("%.", digits, "f (%.", digits, "f)"), m, s)
}


tab_stats = lapply(list(2017:2021, 2022:2024), function(period) {
  # create empty tab with names in first column and NA for groups
  tab = data.frame(
    Name = var_names$varname[match(vars, var_names$var)],
    MeanSD = NA_character_,
    stringsAsFactors = FALSE
  )

  dta2 = dta %>%
    dplyr::filter(as.integer(as.character(year)) %in% period) %>%
    dplyr::mutate(
      pop = pop / 1000,
      N = 1
    )
  stopifnot(!anyNA(dta2[c("soy_price_current", "cattle_price_current")]))
  i = 1
  # fill tab by looping over vars
  for (var in vars) {
    colvals = dta2[[var]]
    # compute per-group formatted strings
    tab$MeanSD[i] = fmt(colvals)
    i = i + 1
  }

  tab = rbind(colnames(tab), c("N", nrow(dta2) / length(period)), tab)
  colnames(tab) = NULL

  format(tab, justify = "left")
})


tab_stat = cbind(
  tab_stats[[1]][, 1:2],
  tab_stats[[2]][, 2]
)
colnames(tab_stat) = NULL
tab_stat[1, ] = c(
  "Variables                             ",
  "2017-2021       ",
  "2022-2024       "
)
print(tab_stat)
```



## Figure 1

```{r Figure1Map, message=F, warning=F, echo=F, fig.width=8, fig.height=9, dpi=150}
# Figure 1: Starlink adoption, mobile coverage, and DETER degradation.
figure1_mobile = readRDS(file.path(
  DATA_DIR, "processed", "mobile_coverage_2021.RDS"
))
deter_map = sf::st_as_sf(readRDS(file.path(
  DATA_DIR, "processed", "deter_map.RDS"
)))
figure1_deter = deter_map %>%
  dplyr::filter(year %in% c(2022L, 2024L))

# Municipal Starlink rates for the two mapped years, from the calendar panel
figure1_starlink = dta %>%
  dplyr::ungroup() %>%
  dplyr::transmute(
    codmun = as.numeric(as.character(codmun)),
    year = as.integer(as.character(year)),
    starlink
  ) %>%
  dplyr::filter(year %in% c(2022L, 2024L))

# This is the panel-matched 772-unit geography.
figure1_mun = readRDS(mun_legal_amazon_geometry) %>%
  sf::st_make_valid() %>%
  sf::st_transform(4674) %>%
  dplyr::mutate(codmun = as.numeric(codmun)) %>%
  dplyr::select(codmun)

figure1_mun = dplyr::bind_rows(
  figure1_mun %>% dplyr::mutate(year = 2022L),
  figure1_mun %>% dplyr::mutate(year = 2024L)
) %>%
  dplyr::left_join(figure1_starlink, by = c("codmun", "year")) %>%
  dplyr::mutate(starlink = tidyr::replace_na(starlink, 0))

# The legend uses quintiles of 2024 municipal adoption, as stated in the paper.
figure1_breaks = stats::quantile(
  figure1_mun$starlink[figure1_mun$year == 2024L],
  probs = seq(0, 1, 0.2),
  na.rm = TRUE,
  names = FALSE
)

figure1_labels = c(
  sprintf("[%.1f - %.1f]", figure1_breaks[1], figure1_breaks[2]),
  sprintf("(%.1f - %.1f]", figure1_breaks[2], figure1_breaks[3]),
  sprintf("(%.1f - %.1f]", figure1_breaks[3], figure1_breaks[4]),
  sprintf("(%.1f - %.1f]", figure1_breaks[4], figure1_breaks[5]),
  sprintf("(%.1f - %.1f]", figure1_breaks[5], figure1_breaks[6])
)

figure1_palette = stats::setNames(
  c("#FFFFFF", "#D8D8D8", "#AEAEAE", "#5F5F5F", "#2B2B2B"),
  figure1_labels
)

figure1_mun = figure1_mun %>%
  dplyr::mutate(
    starlink_bin = cut(
      starlink,
      breaks = figure1_breaks,
      labels = figure1_labels,
      include.lowest = TRUE,
      right = TRUE
    ),
    starlink_bin = factor(starlink_bin, levels = figure1_labels)
  )

make_figure1_panel = function(year_value, panel_label, degr_linewidth) {
  ggplot(figure1_mun %>% dplyr::filter(year == year_value)) +
    geom_sf(
      aes(fill = starlink_bin),
      color = "grey20",
      linewidth = 0.05,
      alpha = 0.8
    ) +
    scale_fill_manual(
      name = "Starlink per 1,000",
      values = figure1_palette,
      limits = figure1_labels,
      drop = FALSE,
      na.value = "#E0E0E0",
      guide = guide_legend(order = 1)
    ) +
    ggnewscale::new_scale_fill() +
    geom_sf(
      data = figure1_mobile,
      aes(fill = "Mobile"),
      alpha = 0.9,
      color = NA
    ) +
    scale_fill_manual(
      name = NULL,
      values = c("Mobile" = "blue"),
      breaks = "Mobile",
      guide = guide_legend(
        order = 2,
        override.aes = list(alpha = 0.9, color = NA)
      )
    ) +
    ggnewscale::new_scale_fill() +
    geom_sf(
      data = figure1_deter %>% dplyr::filter(year == year_value),
      aes(fill = "Degradation"),
      alpha = 0.5,
      color = "red",
      linewidth = degr_linewidth
    ) +
    scale_fill_manual(
      name = NULL,
      values = c("Degradation" = "red"),
      breaks = "Degradation",
      guide = guide_legend(
        order = 3,
        override.aes = list(alpha = 0.9, color = NA)
      )
    ) +
    coord_sf(datum = NA) +
    annotate(
      "text", x = -73, y = 5, label = panel_label,
      size = 3.5, hjust = 0
    ) +
    theme_minimal() +
    theme(
      text = element_text(size = 11),
      legend.text = element_text(size = 10.5),
      legend.title = element_text(size = 11),
      legend.key.height = grid::unit(5.5, "mm"),
      legend.key.width = grid::unit(7, "mm"),
      legend.box.spacing = grid::unit(1, "mm"),
      legend.margin = margin(0, 0, 0, 0),
      plot.margin = margin(0, 0, 0, 0)
    )
}

figure1_panel_2022 = make_figure1_panel(2022L, "(a) 2022", 0.10)
figure1_panel_2024 = make_figure1_panel(2024L, "(b) 2024", 0.05)

figure1_legend = ggpubr::get_legend(
  figure1_panel_2024 + theme(legend.position = "right")
)

figure1_panels = ggpubr::ggarrange(
  figure1_panel_2022 + theme(legend.position = "none"),
  figure1_panel_2024 + theme(legend.position = "none"),
  ncol = 1,
  nrow = 2
)

figure1_map = ggpubr::ggarrange(
  figure1_panels,
  figure1_legend,
  ncol = 2,
  widths = c(1, 0.23)
)

print(figure1_map)
```

## Figure A1

```{r FigureDroughtFireMap, message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=8.5, dpi=150}
figure_a1_years = 2021:2024

# Use the panel-matched 772-unit municipal geometry
figure_a1_mun = readRDS(mun_legal_amazon_geometry) %>%
  sf::st_make_valid() %>%
  dplyr::mutate(codmun = as.numeric(codmun)) %>%
  dplyr::select(codmun)
figure_a1_crs = sf::st_crs(figure_a1_mun)$wkt

figure_a1_metadata = dta %>%
  dplyr::transmute(
    codmun = as.numeric(codmun),
    uf
  ) %>%
  dplyr::distinct()

figure_a1_states = figure_a1_mun %>%
  dplyr::left_join(figure_a1_metadata, by = "codmun") %>%
  dplyr::group_by(uf) %>%
  dplyr::summarise(.groups = "drop") %>%
  sf::st_make_valid()

figure_a1_spei = dta %>%
  dplyr::transmute(
    codmun = as.numeric(codmun),
    year = as.integer(as.character(year)),
    spei12_dec
  ) %>%
  dplyr::filter(year %in% figure_a1_years) %>%
  dplyr::distinct()

figure_a1_drought = figure_a1_mun %>%
  dplyr::left_join(figure_a1_spei, by = "codmun") %>%
  dplyr::mutate(
    spei12_display = pmax(pmin(spei12_dec, 0), -6),
    panel = factor(year, levels = figure_a1_years)
  )

# Compact submission-local inputs reproduce the fixed 2021 coverage layer and
# the 2021--2024 DETER fire-scar polygons without relying on external paths.
figure_a1_mobile = figure1_mobile %>%
  sf::st_make_valid() %>%
  sf::st_transform(figure_a1_crs)
figure_a1_fire = deter_map %>%
  dplyr::filter(alert_class == "fire scar") %>%
  sf::st_transform(figure_a1_crs) %>%
  dplyr::mutate(panel = factor(year, levels = figure_a1_years))
stopifnot(setequal(sort(unique(figure_a1_fire$year)), figure_a1_years))

# Rasterization is for legibility only. The outcome construction continues to
# use the original DETER polygons.
figure_a1_template = terra::rast(
  terra::ext(terra::vect(figure_a1_mun)),
  resolution = 2000,
  crs = figure_a1_crs
)
figure_a1_mobile_raster = terra::rasterize(
  terra::vect(figure_a1_mobile),
  figure_a1_template,
  field = 1,
  background = NA,
  touches = TRUE
)
figure_a1_mobile_cells = terra::as.data.frame(
  figure_a1_mobile_raster,
  xy = TRUE,
  na.rm = TRUE
) %>%
  dplyr::transmute(x, y)

figure_a1_fire_cells = lapply(figure_a1_years, function(year_i) {
  year_raster = terra::rasterize(
    terra::vect(figure_a1_fire %>% dplyr::filter(year == year_i)),
    figure_a1_template,
    field = 1,
    background = NA,
    touches = TRUE
  )

  terra::as.data.frame(year_raster, xy = TRUE, na.rm = TRUE) %>%
    dplyr::transmute(
      x,
      y,
      panel = factor(year_i, levels = figure_a1_years)
    )
}) %>%
  dplyr::bind_rows()

figure_a1_plot = ggplot2::ggplot(figure_a1_drought) +
  ggplot2::geom_sf(ggplot2::aes(fill = spei12_display), color = NA) +
  ggplot2::scale_fill_gradient(
    low = "#d9a400",
    high = "#fffde7",
    limits = c(-6, 0),
    oob = scales::squish,
    breaks = c(-6, -3, 0),
    name = "December SPEI-12 (positive values capped at 0)",
    guide = ggplot2::guide_colorbar(
      order = 1,
      direction = "horizontal",
      title.position = "top",
      title.hjust = 0.5,
      label.position = "bottom",
      barwidth = grid::unit(8, "cm"),
      barheight = grid::unit(0.4, "cm")
    )
  ) +
  ggnewscale::new_scale_fill() +
  ggplot2::geom_tile(
    data = figure_a1_mobile_cells,
    ggplot2::aes(x = x, y = y, fill = "2021 mobile coverage"),
    width = 2000,
    height = 2000,
    color = NA,
    alpha = 0.9,
    inherit.aes = FALSE
  ) +
  ggplot2::geom_tile(
    data = figure_a1_fire_cells,
    ggplot2::aes(x = x, y = y, fill = "DETER fire scar"),
    width = 2600,
    height = 2600,
    color = "#FF0000",
    linewidth = 0.08,
    alpha = 1,
    inherit.aes = FALSE
  ) +
  ggplot2::scale_fill_manual(
    values = c(
      "2021 mobile coverage" = "#0000FF",
      "DETER fire scar" = "#FF0000"
    ),
    name = NULL,
    breaks = c("2021 mobile coverage", "DETER fire scar"),
    guide = ggplot2::guide_legend(
      order = 2,
      direction = "horizontal",
      nrow = 1,
      byrow = TRUE,
      override.aes = list(alpha = 1)
    )
  ) +
  ggplot2::geom_sf(
    data = figure_a1_states,
    fill = NA,
    color = "grey25",
    linewidth = 0.25,
    inherit.aes = FALSE
  ) +
  ggplot2::facet_wrap(~panel, ncol = 2, drop = FALSE) +
  ggplot2::coord_sf(expand = FALSE, datum = NA) +
  ggplot2::theme_void(base_size = 11) +
  ggplot2::theme(
    strip.text = ggplot2::element_text(face = "bold", size = 11),
    legend.position = "top",
    legend.box = "vertical",
    legend.justification = "center",
    plot.margin = ggplot2::margin(5.5, 5.5, 5.5, 5.5)
  )

print(figure_a1_plot)
```



## Estimations

```{r Covariates, echo=F}
varlist = var_names$var[
  !grepl("_prod", var_names$var, ignore.case = TRUE) &
    var_names$type == "out"
]
varlist_missing = setdiff(varlist, names(dta))
if (length(varlist_missing)) {
  stop("Missing outcomes: ", paste(varlist_missing, collapse = ", "))
}

# Outcomes and display metadata shared by the calendar-year diagnostics and
# appendix heterogeneity checks.
main_outcome_specs = tibble::tribble(
  ~outcome,                       ~label,                         ~ylab,
  "r_n_mun_deter_degr",          "Degradation flags",           "flags per 100 km\u00b2",
  "r_area_mun_deter_degr",       "Degradation area",            "% 2021 forest",
  "r_area_ndpf_deter_degr",      "Degradation area in UPF",     "% 2021 forest",
  "rm_n_igni",                   "Ignitions",                   "ignitions per 100 km\u00b2 munic. area",
  "disp_shhiunb",                "Ignition spatial HHI",        "index",
  "sh21_agr_pec",                "Forest transition to pasture", "% 2021 forest",
  "pm1mun",                      "PM\u2081",                         "\u00b5g/m\u00b3",
  "r_n_mun_police",              "Enforcement fines",           "fines per 100 km\u00b2"
)
main_outcomes = main_outcome_specs$outcome
main_outcome_labels = setNames(main_outcome_specs$label, main_outcomes)
main_outcome_ylabs = setNames(main_outcome_specs$ylab, main_outcomes)
main_outcome_titles = setNames(
  paste0("(", letters[seq_along(main_outcomes)], ") ", main_outcome_specs$label),
  main_outcomes
)

heterogeneity_label_overrides = c(
  r_n_mun_deter_degr = "Degradation flags (per 100 km\u00b2 of 2021 forest)",
  r_area_mun_deter_degr = "Degradation area (% of 2021 forest)",
  r_area_ndpf_deter_degr = "  in UPF",
  r_area_mun_prodes_defo = "PRODES deforestation (% of 2021 forest)",
  r_area_mun_prodes_defo_dpdp = "  PRODES deforestation by progressive degradation (% of 2021 forest)",
  rm_n_igni = "Ignitions (per 100 km\u00b2 munic. area)",
  disp_shhiunb = "Ignition spatial HHI (index)",
  sh21_agr_pec = "Forest transition to pasture (% of 2021 forest)",
  pm1mun = "PM\u2081 (\u03bcg/m\u00b3)",
  r_n_mun_police = "Enforcement fines (per 100 km\u00b2 of 2021 forest)"
)

heterogeneity_label = function(y) {
  if (y %in% names(heterogeneity_label_overrides)) {
    return(unname(heterogeneity_label_overrides[y]))
  }
  var_names$varname[match(y, var_names$var)]
}

# The manuscript reports Model I without controls and Model II with the fixed
# control set.
covlist = list("1", model_controls)


```

```{r EstimationsSetupSE, message=F, warning=F, echo=F}
# Estimation data for both reported models.
d_se = dta %>% filter(reg_sample == 1)

# `MOPtest()` requires a momentfit model and cannot absorb `fixest` fixed effects.
# Residualize the outcome, endogenous regressor, and instruments jointly using
# the reported controls and fixed effects, then let momentfit compute the test.
mop_test = function(data, yvar, controls, iv) {
  variables = c(yvar, "starlink", iv)
  required = unique(c(
    variables,
    all.vars(stats::as.formula(paste("~", paste(controls, collapse = " + ")))),
    "codmun", "uf", "year", "microreg"
  ))
  data_use = droplevels(data[stats::complete.cases(data[required]), ])

  residual_fits = fixest::feols(
    stats::as.formula(paste0(
      "c(", paste(variables, collapse = ", "), ") ~ ",
      paste(controls, collapse = " + "),
      " | codmun + interaction(uf, year)"
    )),
    data = data_use,
    notes = FALSE,
    nthreads = 1
  )

  z_names = paste0("z", seq_along(iv))
  mop_data = as.data.frame(lapply(residual_fits, stats::residuals))
  names(mop_data) = c("mop_y", "mop_d", z_names)
  mop_data$cluster = data_use$microreg[fixest::obs(residual_fits[[1]])]

  momentfit::MOPtest(
    momentfit::momentModel(
      mop_y ~ mop_d - 1,
      stats::reformulate(z_names, intercept = FALSE),
      data = mop_data,
      vcov = "CL",
      vcovOptions = list(cluster = ~cluster)
    ),
    tau = 0.10,
    size = 0.05,
    estMethod = "TSLS",
    simplified = TRUE,
    print = FALSE
  )
}
```


- Model I: no covariates
- Model II: all covariates

```{r Estimations, message=F, warning=F, R.options = list(width = 1200), echo=F}
# Fit models
est_fits = vector("list", length(covlist))

bd_fig = data.frame() # Store results to create figures

for (m in seq_along(covlist)) {
  est_fits[[m]] = list()
  for (y in varlist) {
    model = paste0(
      y, " ~ ", paste0(covlist[[m]], collapse = " + "), # Covariates
      " | codmun + interaction(uf, year)", # fixed-effects
      " | starlink ~ coverarea:year"
    ) %>% as.formula()

    est = feols(model, data = d_se, cluster = ~microreg, notes = FALSE)
    est_fits[[m]][[y]] = est

    # First-stage and figure data come from the full model (last in covlist).
    if (m == length(covlist)) {
      d_op = dplyr::filter(d_se, !is.na(.data[[y]]))
      op_F_y = mop_test(d_op, y, covlist[[m]], calendar_iv_names)[["Feff"]]

      aux = broom::tidy(est, conf.int = TRUE, conf.level = 0.95) %>%
        rename(conf.low95 = conf.low, conf.high95 = conf.high) %>%
        left_join(
          broom::tidy(est, conf.int = TRUE, conf.level = 0.90) %>%
            select(term, conf.low, conf.high) %>%
            rename(conf.low90 = conf.low, conf.high90 = conf.high),
          by = "term"
        ) %>%
        filter(term == "fit_starlink") %>%
        mutate(
          term = y,
          N = nobs(est),
          op_F = op_F_y,
          ymean = mean(d_se[[y]], na.rm = TRUE),
          n_micro = dplyr::n_distinct(d_se$microreg[!is.na(d_se[[y]])]),
          n_munic = dplyr::n_distinct(d_se$codmun[!is.na(d_se[[y]])])
        )

      # Saving results for figures
      bd_fig = bind_rows(bd_fig, aux)
    }
  }
}

# Downstream chunks use the full covariate set.
print(paste0("Covariates (full model): ", paste(model_controls, collapse = ", ")))

# Build the estimation table with one column per model. N is reported by outcome
# because tenure and ignition-concentration samples differ.
result = data.frame(matrix(
  ncol = 1 + length(covlist),
  nrow = 2 + 2 * length(varlist) + 1,
  data = ""
))
result[1, 1 + ceiling(length(covlist) / 2)] = "Starlink"
result[2, 1] = "Outcomes"
result[2, 2:ncol(result)] = paste0("  (", as.roman(seq_along(covlist)), ")")

result = cbind(result, data.frame(ci95 = "", t = "", p = "", N = "", opF = ""))
result[2, (ncol(result) - 4):ncol(result)] = c("95% CI", "t-stat", "p-val", "N", "OP F")
result[3 + 2 * length(varlist), ] = "-------"

i = 3
for (vardep in varlist) {
  rownames(result)[i] = paste0(vardep, "_est")
  rownames(result)[i + 1] = paste0(vardep, "_se")
  result[i, 1] = var_names$varname[which(vardep == var_names$var)]
  i = i + 2
}

for (m in seq_along(covlist)) {
  for (y in varlist) {
    est = est_fits[[m]][[y]]
    if (!("fit_starlink" %in% rownames(est$coeftable))) {
      result[paste0(y, "_est"), m + 1] = "---"
      next
    }
    ct = est$coeftable
    b = ct["fit_starlink", 1]
    se = ct["fit_starlink", 2]
    tval = ct["fit_starlink", 3]
    p = ct["fit_starlink", 4]

    result[paste0(y, "_est"), m + 1] = paste0(
      sprintf("%.3f", b),
      case_when(p <= .01 ~ "***", p <= .05 ~ "**", p <= .10 ~ "*", TRUE ~ "")
    )
    result[paste0(y, "_se"), m + 1] = paste0("(", sprintf("%.3f", se), ")")

    if (m == length(covlist)) {
      ci = as.numeric(confint(est, "fit_starlink", level = 0.95)[1, ])
      result[paste0(y, "_est"), "ci95"] = paste0(
        "[", sprintf("%.3f", ci[1]), ", ", sprintf("%.3f", ci[2]), "]"
      )
      result[paste0(y, "_est"), "t"] = sprintf("%.3f", tval)
      result[paste0(y, "_est"), "p"] = sprintf("%.3f", p)
      result[paste0(y, "_est"), "N"] = formatC(nobs(est), format = "d", big.mark = ",")
      op_f = bd_fig$op_F[match(y, bd_fig$term)]
      result[paste0(y, "_est"), "opF"] = if (is.na(op_f)) "." else sprintf("%.1f", op_f)
    }
  }
}
rownames(result) = seq_len(nrow(result))
colnames(result) = NULL

print(format(
  result[, -c((ncol(result) - 4):(ncol(result) - 3))],
  justify = "left"
))
```


## Diagnostics
```{r DiagnosticsSetup, include=FALSE}
outcome_label = function(x) {
  label = var_names$varname[match(x, var_names$var)]
  ifelse(is.na(label), x, label)
}

fit_starlink = function(
  data, outcome, iv_rhs = "coverarea:year", controls = model_controls
) {
  fixest::feols(
    stats::as.formula(paste0(
      outcome, " ~ ", paste(controls, collapse = " + "),
      " | codmun + interaction(uf, year) | starlink ~ ", iv_rhs
    )),
    data = dplyr::filter(data, reg_sample == 1),
    cluster = ~microreg,
    notes = FALSE,
    nthreads = 1
  )
}

estimate_starlink = function(
  data, outcome, iv_rhs = "coverarea:year", controls = model_controls
) {
  est = fit_starlink(data, outcome, iv_rhs, controls)
  tab = broom::tidy(est) %>% filter(term == "fit_starlink")
  ci95 = as.numeric(confint(est, "fit_starlink", level = 0.95)[1, ])
  ci90 = as.numeric(confint(est, "fit_starlink", level = 0.90)[1, ])

  tibble(
    outcome = outcome,
    estimate = tab$estimate,
    std.error = tab$std.error,
    statistic = tab$statistic,
    p.value = tab$p.value,
    conf.low95 = ci95[1],
    conf.high95 = ci95[2],
    conf.low90 = ci90[1],
    conf.high90 = ci90[2],
    N = stats::nobs(est)
  )
}

robust_full = map_dfr(main_outcomes, ~ estimate_starlink(dta_normal, .x)) %>%
  mutate(sample = "Full")

# Per-municipality first-stage coefficient-leverage share.
municipality_leverage = function(panel) {
  required = c(
    "starlink", "coverarea", "codmun", "uf", "year",
    all.vars(stats::as.formula(paste("~", paste(model_controls, collapse = " + "))))
  )
  d = panel %>%
    dplyr::filter(reg_sample == 1) %>%
    tidyr::drop_na(dplyr::all_of(required))
  instruments = grep("^ca_\\d+$", names(d), value = TRUE)
  instruments = instruments[vapply(
    instruments,
    function(x) !all(is.na(d[[x]])) && stats::sd(d[[x]], na.rm = TRUE) > 0,
    logical(1)
  )]
  d = tidyr::drop_na(d, dplyr::all_of(instruments))

  fs = fixest::feols(
    stats::as.formula(paste0(
      "starlink ~ ",
      paste(c(instruments, model_controls), collapse = " + "),
      " | codmun + interaction(uf, year)"
    )),
    data = d,
    notes = FALSE
  )
  instrument_coef = stats::coef(fs)[instruments]
  instruments = instruments[!is.na(instrument_coef)]
  instrument_coef = instrument_coef[!is.na(instrument_coef)]
  instrument_fits = fixest::feols(
    stats::as.formula(paste0(
      "c(", paste(instruments, collapse = ", "), ") ~ ",
      paste(model_controls, collapse = " + "),
      " | codmun + interaction(uf, year)"
    )),
    data = d,
    notes = FALSE
  )
  d = d[fixest::obs(instrument_fits[[1]]), ]
  instrument_residuals = do.call(cbind, lapply(instrument_fits, stats::residuals))
  d$leverage_mass = as.numeric(instrument_residuals %*% instrument_coef)^2

  d %>%
    dplyr::group_by(codmun) %>%
    dplyr::summarise(
      leverage_mass = sum(leverage_mass, na.rm = TRUE), .groups = "drop"
    ) %>%
    dplyr::mutate(leverage_share = leverage_mass / sum(leverage_mass))
}
```


```{r WeakIVSetup, message=F, warning=F, echo=F}
# Shared microregion-clustered OP effective F helpers for the weak-IV diagnostics.
split_controls_for_ivdiag = function(controls, data) {
  is_bare = controls %in% names(data)
  list(formula = controls[!is_bare], bare = controls[is_bare])
}

# Olea-Pflueger effective F with the manuscript's microregion clustering.
op_F_clustered = function(
  data, yvar, formula_controls, bare_controls, iv
) {
  mop_test(
    data = data,
    yvar = yvar,
    controls = c(formula_controls, bare_controls),
    iv = iv
  )[["Feff"]]
}
```


### Exogeneity

#### Event-study placebo

The event studies use 2022 as the omitted year. The table reports joint Wald p-values for all pre-2022 coefficients.


```{r PlaceboTestMicroreg, message=FALSE, warning=FALSE, fig.height=8, fig.width=10, echo=FALSE}
# Calendar-year event-study placebo with standard errors clustered by IBGE 2017 microregion
placebo_outcomes = main_outcomes

placebo_pretrend_test = function(model, years = 2017:2021) {
  keep_regex = paste0("^ca_(", paste(years, collapse = "|"), ")$")
  test = tryCatch(
    fixest::wald(model, keep = keep_regex, print = FALSE),
    error = function(e) NULL
  )
  if (is.null(test)) {
    return(tibble(pretrend_F = NA_real_, pretrend_p = NA_real_))
  }
  tibble(
    pretrend_F = as.numeric(test$stat),
    pretrend_p = as.numeric(test$p)
  )
}

placebo_pretrend_calendar = tibble()
placebo_panels = list()

# Placebo panel: full dta, including the pre-period leads.
d_plm = dta %>% mutate(codmun = as.numeric(codmun))

for (var in placebo_outcomes) {
  placebo_start_year = if (
    var %in% c('rm_n_igni', 'disp_shhiunb')
  ) 2019L else 2017L

  placebo_years = placebo_start_year:2024

  ca_terms_var = paste(
    paste0('ca_', setdiff(placebo_years, 2022)), collapse = ' + '
  )
  est_data = d_plm %>% filter(year %in% placebo_years)

  model = paste0(
    var, " ~ ", ca_terms_var,  " + ", paste(model_controls, collapse = " + "),
    " | codmun + interaction(uf, year)") %>%
    as.formula()

  est = feols(model, data = est_data, cluster = ~microreg, notes = FALSE)

  placebo_pretrend_calendar = bind_rows(
    placebo_pretrend_calendar,
    bind_cols(
      tibble(outcome = var),
      placebo_pretrend_test(est, years = placebo_start_year:2021)
    )
  )

  ct = est$coeftable
  trm = rownames(ct)
  ci95 = confint(est, level = 0.95)
  ci90 = confint(est, level = 0.90)

  tab = tibble(
    term = trm,
    estimate = ct[, 1],
    conf.low95 = ci95[trm, 1],
    conf.high95 = ci95[trm, 2],
    conf.low90 = ci90[trm, 1],
    conf.high90 = ci90[trm, 2]
  ) %>%
    filter(grepl("^ca_", term)) %>%
    mutate(year = as.numeric(gsub("ca_", "", term))) %>%
    filter(!is.na(year)) %>%
    select(-term) %>%
    bind_rows(tibble(
      year = 2022,
      estimate = 0,
      conf.low95 = 0,
      conf.high95 = 0,
      conf.low90 = 0,
      conf.high90 = 0
    )) %>%
    arrange(year)

  var_name = trimws(var_names$varname[which(var == var_names$var)])

  placebo_panels[[var]] = ggplot(tab, aes(x = year, y = estimate)) +
      geom_hline(
        yintercept = 0,
        color = "black",
        size = 0.5,
        linetype = "dashed"
      ) +
      geom_vline(
        xintercept = 2022,
        color = "black",
        size = 0.5,
        linetype = "dashed"
      ) +
      geom_linerange(
        aes(ymin = conf.low95, ymax = conf.high95, colour = "95% CI"),
        position = position_dodge(width = 0.5),
        size = 2,
        show.legend = TRUE,
        key_glyph = "vline"
      ) +
      scale_colour_manual(
        name = NULL,
        breaks = "95% CI",
        values = c("95% CI" = "grey80"),
        guide = guide_legend(
          order = 2,
          keyheight = grid::unit(0.9, "cm"),
          override.aes = list(shape = NA, linetype = 1, linewidth = 3)
        )
      ) +
      ggnewscale::new_scale_color() +
      geom_point(
        position = position_dodge(width = 0.5),
        size = 1.5,
        color = "red"
      ) +
      labs(
        title = var_name,
        x = "Year",
        y = "Estimate",
        color = "Group"
      ) +
      theme_minimal() +
      theme(
        text = element_text(size = 10),
        plot.title = element_text(size = 11.25),
        panel.grid.minor.y = element_blank(),
        legend.key.height = grid::unit(0.5, "cm"),
        legend.spacing.y = grid::unit(0.1, "cm")
      )
}

# Display the selected outcomes as one labeled grid in the Exogeneity section
selected_placebo_panels = lapply(
  seq_along(main_outcomes),
  function(i) {
    oc = main_outcomes[i]
    stopifnot(!is.null(placebo_panels[[oc]]))
    pretrend_p = placebo_pretrend_calendar %>%
      filter(outcome == oc) %>%
      pull(pretrend_p)
    stopifnot(length(pretrend_p) == 1, !is.na(pretrend_p))
    pretrend_label = sprintf('%.3f', pretrend_p)
    panel_title = if (identical(oc, "pm1mun")) {
      bquote(paste("(", .(letters[i]), ") ", PM[1]))
    } else {
      main_outcome_titles[[oc]]
    }
    if (identical(oc, 'r_area_over_ndpf_large_deter_degr')) {
      panel_title = sub(' claims$', '', panel_title)
    }
    panel_title = if (identical(oc, 'pm1mun')) {
      bquote(
        paste(
          '(', .(letters[i]), ') ', PM[1],
          ' (pre-trends p = ', .(pretrend_label), ')'
        )
      )
    } else {
      paste0(panel_title, ' (pre-trends p = ', pretrend_label, ')')
    }
    placebo_panels[[oc]] +
      labs(
        title = panel_title,
        x = NULL,
        y = main_outcome_ylabs[[oc]]
      ) +
      theme(legend.position = "bottom")
  }
)

fig_placebo_selected_8 = ggpubr::ggarrange(
  plotlist = selected_placebo_panels,
  ncol = 2,
  nrow = 4,
  common.legend = TRUE,
  legend = "bottom",
  align = "hv"
)
print(fig_placebo_selected_8)
```



## Robustness

### Inference
```{r InferenceAR, message=F, warning=F, R.options = list(width = 1200), echo=F}
# Anderson-Rubin test + inverted 95% confidence set via ivDiag::AR_test()
ar_ci_ivdiag = function(
  data,
  yvar,
  formula_controls,
  bare_controls,
  iv,
  fe_cols = c("codmun", "uf_year"),
  cluster_col = "microreg",
  endog = "starlink",
  alpha = 0.05
) {
  if (!requireNamespace("ivDiag", quietly = TRUE)) {
    return(NULL)
  }
  data_aug = as.data.frame(data)
  if ("uf_year" %in% fe_cols && !"uf_year" %in% names(data_aug)) {
    data_aug$uf_year = interaction(data_aug$uf, data_aug$year, drop = TRUE)
  }
  mm_cols = character(0)
  if (length(formula_controls) > 0) {
    mm_form = stats::as.formula(paste(
      "~ -1 +",
      paste(formula_controls, collapse = " + ")
    ))
    mm = stats::model.matrix(
      mm_form,
      data = data_aug,
      na.action = stats::na.pass
    )
    colnames(mm) = make.names(colnames(mm), unique = TRUE)
    mm_cols = colnames(mm)
    data_aug = cbind(data_aug, as.data.frame(mm))
  }
  suppressWarnings(tryCatch(
    ivDiag::AR_test(
      data = data_aug,
      Y = yvar,
      D = endog,
      Z = iv,
      controls = c(bare_controls, mm_cols),
      FE = fe_cols,
      cl = cluster_col,
      CI = TRUE,
      alpha = alpha,
      prec = 6
    ),
    error = function(e) NULL
  ))
}

# One row per outcome: 2SLS estimate / SE / p (estimate_starlink, the same
# wrapper the other robustness checks use) alongside AR p and the AR 95% set.
inference_table = function(
  data,
  outcomes,
  iv = calendar_iv_names,
  iv_rhs = paste(iv, collapse = " + ")
) {
  purrr::map_dfr(outcomes, function(y) {
    # Instruments with no within-sample variation are removed outcome by outcome.
    d_y = dplyr::filter(data, reg_sample == 1, !is.na(.data[[y]]))
    iv_y = intersect(iv, names(d_y))
    iv_y = iv_y[vapply(iv_y, function(z) {
      vz = d_y[[z]]
      any(is.finite(vz)) && stats::sd(vz, na.rm = TRUE) > 0
    }, logical(1))]
    if (!length(iv_y)) {
      return(tibble::tibble(
        Outcome = outcome_label(y), Estimate = NA_real_, SE = NA_real_,
        p = NA_real_, `AR F` = NA_real_, `AR p` = NA_real_,
        `OP F` = NA_real_, `AR 95% CI` = NA_character_, Bounded = NA,
        N = nrow(d_y)
      ))
    }
    iv_rhs_y = paste(iv_y, collapse = " + ")
    est = estimate_starlink(data, y, iv_rhs = iv_rhs_y)
    ivd_split = split_controls_for_ivdiag(model_controls, d_y)
    op_F = tryCatch(
      op_F_clustered(
        data = d_y,
        yvar = y,
        formula_controls = ivd_split$formula,
        bare_controls = ivd_split$bare,
        iv = iv_y
      ),
      error = function(e) NA_real_
    )
    ar = ar_ci_ivdiag(
      data = d_y,
      yvar = y,
      formula_controls = ivd_split$formula,
      bare_controls = ivd_split$bare,
      iv = iv_y
    )
    # AR_test returns Fstat = c(F, df1, df2, p) and ci.print as ONE formatted
    # string, which already carries the union / infinite-bound cases.
    ar_p = if (!is.null(ar)) unname(ar$Fstat["p"]) else NA_real_
    ar_F = if (!is.null(ar)) unname(ar$Fstat["F"]) else NA_real_
    tibble::tibble(
      Outcome = outcome_label(y),
      Estimate = est$estimate,
      SE = est$std.error,
      p = est$p.value,
      `AR F` = ar_F,
      `AR p` = ar_p,
      `OP F` = op_F,
      `AR 95% CI` = if (is.null(ar)) NA_character_ else ar$ci.print,
      Bounded = if (is.null(ar)) NA else isTRUE(ar$bounded),
      N = est$N
    )
  })
}
# The same nine outcomes are used in the AR and outcome-normalization checks.
inference_specs = tibble::tribble(
  ~outcome,                                  ~label,
  "r_n_mun_deter_degr",                     "Degradation (flags per 100 km2)",
  "r_area_mun_deter_degr",                  "Degradation area",
  "r_area_ndpf_deter_degr",                 "  UPF",
  "r_area_over_ndpf_large_deter_degr",      "    with large-property claims",
  "rm_n_igni",                               "Ignitions (per 100 km2 munic. area)",
  "disp_shhiunb",                            "Ignition spatial HHI",
  "sh21_agr_pec",                            "Forest transition to pasture",
  "pm1mun",                                  "PM1",
  "r_n_mun_police",                         "Enforcement fines (per 100 km2)"
)

inference_outcomes = inference_specs$outcome
missing_inference = setdiff(inference_outcomes, names(dta_normal))

if (length(missing_inference)) {
  stop("Missing requested AR outcomes: ", paste(missing_inference, collapse = ", "))
}

inference_normal = inference_table(dta_normal, inference_outcomes)
inference_normal = inference_normal %>%
  dplyr::mutate(
    Outcome = inference_specs$label[match(Outcome, outcome_label(inference_specs$outcome))]
  )

print(
  as.data.frame(
    dplyr::select(
      inference_normal,
      Outcome, Estimate, SE, p, "AR p", "AR 95% CI", "OP F", N
    )
  ),
  row.names = FALSE,
  digits = 4
)

```


### Outcome normalization

```{r NormalizationRobustness, message=F, warning=F, R.options = list(width = 1200), echo=F}
# Outcome-normalization robustness
nz_dat = dta_normal %>%
  dplyr::mutate(nz_pasture_area = (sh21_agr_pec / 100) * area_forest * 1000)

# Raw numerators and denominator-based versions displayed in the table.
nz_specs = tibble::tribble(
 ~raw, ~forest, ~municipal, ~label,
 "n_mun_deter_degr", "r_n_mun_deter_degr", "rm_n_mun_deter_degr", "Degradation flags",
 "area_mun_deter_degr", "r_area_mun_deter_degr", "rm_area_mun_deter_degr", "Degradation area",
 "area_ndpf_deter_degr", "r_area_ndpf_deter_degr", "rm_area_ndpf_deter_degr", " UPF",
 "area_over_ndpf_large_deter_degr", "r_area_over_ndpf_large_deter_degr", "rm_area_over_ndpf_large_deter_degr", " with large-property claims",
 "n_igni", "r_n_igni", "rm_n_igni", "Ignitions",
 "disp_shhiunb", NA_character_, "disp_shhiunb", "Ignition spatial HHI\u2020",
 "nz_pasture_area", "sh21_agr_pec", "shm_agr_pec", "Forest transition to pasture",
 "pm1mun", NA_character_, "pm1mun", "PM1\u2020",
 "n_mun_police", "r_n_mun_police", "rm_n_mun_police", "Enforcement fines"
)


# Fixed log constants by measurement unit: 1 for counts, 0.001 km2 for
# raw areas, 0.01 for the HHI, and 0 for strictly positive PM1.
nz_logc = c(
  n_mun_deter_degr = 1,
  area_mun_deter_degr = 0.001,
  area_ndpf_deter_degr = 0.001,
  area_over_ndpf_large_deter_degr = 0.001,
  n_igni = 1,
  disp_shhiunb = 0.01,
  nz_pasture_area = 0.001,
  pm1mun = 0,
  n_mun_police = 1
)

nz_forms = tibble::tribble(
  ~tag,  ~label,
  "for", "2021 forest area",
  "mun", "Municipal area",
  "ihs", "IHS(y)",
  "log", "log(y + c)"
)

nz_rows = list()
for (i in seq_len(nrow(nz_specs))) {
  for (j in seq_len(nrow(nz_forms))) {
    raw = nz_specs$raw[i]
    tag = nz_forms$tag[j]
    source_var = switch(
      tag,
      "for" = nz_specs$forest[i],
      "mun" = nz_specs$municipal[i],
      "ihs" = raw,
      "log" = raw
    )

    cell = "\u2014"
    if (!is.na(source_var)) {
      nz_dat$nz_outcome = switch(
        tag,
        "for" = nz_dat[[source_var]],
        "mun" = nz_dat[[source_var]],
        "ihs" = asinh(nz_dat[[raw]]),
        "log" = log(nz_dat[[raw]] + nz_logc[[raw]])
      )
      estimate = broom::tidy(fit_starlink(nz_dat, "nz_outcome")) %>%
        dplyr::filter(term == "fit_starlink")
      stars = dplyr::case_when(
        estimate$p.value <= 0.01 ~ "***",
        estimate$p.value <= 0.05 ~ "**",
        estimate$p.value <= 0.10 ~ "*",
        TRUE ~ ""
      )
      cell = sprintf("%.4g%s", estimate$estimate, stars)
    }

    nz_rows[[length(nz_rows) + 1]] = tibble::tibble(
      outcome = nz_specs$label[i], form = nz_forms$label[j], cell
    )
  }
}

knitr::kable(
  dplyr::bind_rows(nz_rows) %>%
    dplyr::mutate(
      outcome = factor(outcome, levels = nz_specs$label),
      form = factor(form, levels = nz_forms$label)
    ) %>%
    dplyr::arrange(outcome, form) %>%
    dplyr::select(Outcome = outcome, Transformation = form, cell) %>%
    tidyr::pivot_wider(names_from = Transformation, values_from = cell),
  align = paste0("l", strrep("r", nrow(nz_forms))),
  caption = paste0(
    "Outcome normalization: coefficient on instrumented ",
    "Starlink, microregion-clustered. Transformations are not on a ",
    "common unit; compare sign and significance across each outcome row."
  )
)

```


### Coefficient-leverage deletion

**Identifying-mass map.** Each municipality is shaded by quintile of its coefficient-leverage share in the instrument-predicted Starlink component.

```{r MapVarShareMass, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=4.3, dpi=150}
# Identifying-mass map =====================================================
suppressPackageStartupMessages(library(sf))

# Per-municipality coefficient-leverage share
muni_influence = municipality_leverage(dta_normal) %>%
  mutate(leverage_share_pct = leverage_share * 100)

# Panel-matched Legal Amazon municipality polygons.
iv_mass_poly = readRDS(mun_legal_amazon_geometry) %>%
  st_make_valid() %>%
  st_transform(4674) %>%
  st_make_valid() %>%
  mutate(codmun = as.numeric(codmun)) %>%
  select(codmun)

# Mojui dos Campos (1504752) was split off Santarem (1506807) in 2013
iv_mass_poly = iv_mass_poly %>%
  mutate(codmun = ifelse(codmun == 1504752, 1506807, codmun)) %>%
  group_by(codmun) %>%
  summarise(.groups = "drop") %>%
  st_make_valid()

# Map builder
make_iv_mass_map = function(share_df, poly = iv_mass_poly) {
  # Leverage share as a percentage of total identifying variation.
  df = poly %>%
    left_join(
      share_df %>% select(codmun, leverage_share_pct),
      by = "codmun"
    )

  brks = quantile(df$leverage_share_pct, probs = seq(0, 1, 0.2), na.rm = TRUE)
  brks = unique(brks)
  if (length(brks) < 2) {
    return(NULL)
  }
  labs_q = sprintf("[%.3f - %.3f]", brks[-length(brks)], brks[-1])
  df = df %>%
    mutate(
      leverage_share_q = cut(
        leverage_share_pct,
        breaks = brks,
        labels = labs_q,
        include.lowest = TRUE
      )
    )

  # State (UF) outlines: dissolve the municipalities by their 2-digit IBGE prefix.
  states = df %>%
    mutate(uf = codmun %/% 100000L) %>%
    group_by(uf) %>%
    summarise(.groups = "drop")

  ggplot() +
    geom_sf(data = df, fill = "grey90", color = "grey80", linewidth = 0.05) +
    geom_sf(
      data = dplyr::filter(df, !is.na(leverage_share_q)),
      aes(fill = leverage_share_q),
      color = "grey80",
      linewidth = 0.05
    ) +
    geom_sf(data = states, fill = NA, color = "black", linewidth = 0.3) +
    scale_fill_brewer(
      palette = "OrRd",
      drop = FALSE,
      name = "Leverage-share quintile\n(% of total identifying variation)"
    ) +
    coord_sf(expand = FALSE) +
    theme_void(base_size = 11) +
    theme(
      legend.position = "right",
      legend.key.size = unit(0.9, "lines"),
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8),
      plot.margin = margin(2, 2, 2, 2)
    )
}


print(make_iv_mass_map(muni_influence))
```


```{r LeverageShareHist, include=FALSE, message=FALSE, warning=FALSE}
# Inputs for the manuscript coefficient-leverage deletion figure.

# Full sample followed by 0.5-percentage-point upper-tail deletions through 2%.
leverage_trim_cuts = tibble(
  tail = seq(0, 0.02, by = 0.005),
  sample = c("Full", paste0("Top ", 100 * tail[-1], "%"))
) %>%
  mutate(
    max_retained_pct = quantile(
      muni_influence$leverage_share_pct, probs = 1 - tail, na.rm = TRUE
    )
  )

# Histogram with all four requested upper-tail cutoffs marked, using the same
# colors as the trimmed-estimate panels. The per-municipality
# leverage share spans several orders of magnitude (a long thin upper tail over
# a mass of near-zero shares), so the x-axis is shown on a log10 scale. Zero
# shares cannot be displayed on a logarithmic axis and are omitted.
hist_df = muni_influence %>%
  dplyr::filter(leverage_share_pct > 0)

# Use preliminary equal-width log10 bins to detect an isolated sparse lower
# tail. Begin at the first run of three occupied bins; if no such run exists,
# retain the full positive support.
log_hist_breaks_all = 10^seq(
  log10(min(hist_df$leverage_share_pct, na.rm = TRUE)),
  log10(max(hist_df$leverage_share_pct, na.rm = TRUE)),
  length.out = 41
)
log_hist_counts_all = hist(
  hist_df$leverage_share_pct,
  breaks = log_hist_breaks_all,
  plot = FALSE
)$counts
occupied_run_start = which(
  head(log_hist_counts_all, -2) > 0 &
    log_hist_counts_all[2:(length(log_hist_counts_all) - 1)] > 0 &
    tail(log_hist_counts_all, -2) > 0
)
first_hist_bin = if (length(occupied_run_start) > 0) occupied_run_start[1] else 1L
hist_x_min = log_hist_breaks_all[first_hist_bin]
hist_df = hist_df %>%
  dplyr::filter(leverage_share_pct >= hist_x_min)

# Recompute 40 equal-width log10 bins over the displayed support.
log_hist_breaks = 10^seq(
  log10(hist_x_min),
  log10(max(hist_df$leverage_share_pct, na.rm = TRUE)),
  length.out = 41
)
# Shared row order and grayscale key for the manuscript composite figure. Full
# is red and appears first; increasingly stringent trims run downward in
# progressively darker gray. The same shades mark the histogram cutoffs.
trim_panel_levels = leverage_trim_cuts$sample
trim_panel_cols = c(
  "Full" = "#e41a1c",
  "Top 0.5%" = "grey72",
  "Top 1%" = "grey58",
  "Top 1.5%" = "grey44",
  "Top 2%" = "grey30"
)
fig_leverage_hist = ggplot(hist_df, aes(x = leverage_share_pct)) +
  geom_histogram(
    breaks = log_hist_breaks,
    fill = "red",
    alpha = 0.70,
    color = "white",
    linewidth = 0.2
  ) +
  geom_vline(
    data = dplyr::filter(leverage_trim_cuts, tail > 0),
    aes(xintercept = max_retained_pct, color = sample),
    linewidth = 0.35,
    linetype = "dashed"
  ) +
  scale_color_manual(values = trim_panel_cols[trim_panel_levels[-1]]) +
  scale_x_log10(
    limits = range(log_hist_breaks),
    expand = ggplot2::expansion(mult = c(0.005, 0.005)),
    breaks = c(0.001, 0.003, 0.01, 0.03, 0.1, 0.3, 1, 3, 10),
    labels = function(b) {
      sub(
        "^0\\.",
        ".",
        trimws(format(b, drop0trailing = TRUE, scientific = FALSE))
      )
    }
  )

```

```{r RobustTrimLeverage, message=FALSE, warning=FALSE, echo=FALSE, fig.height=6.3, fig.width=10, dpi=150}
# Coefficient-leverage deletion sensitivity --------------------------------

# One-sided (upper-tail) trim (most identifying leverage)
make_leverage_trim_data = function(cutoff) {
  keep_mun = muni_influence %>%
    filter(leverage_share_pct <= cutoff) %>%
    pull(codmun)

  dta_normal %>% filter(codmun %in% keep_mun)
}

leverage_trim_est = bind_rows(
  robust_full,
  map_dfr(which(leverage_trim_cuts$tail > 0), function(i) {
    data_i = make_leverage_trim_data(leverage_trim_cuts$max_retained_pct[i])
    map_dfr(main_outcomes, ~ estimate_starlink(data_i, .x)) %>%
      mutate(sample = leverage_trim_cuts$sample[i])
  })
)

# === Paper panel: 3x3 grid (8 outcomes + leverage-share histogram) ===========
# The eight coefficient panels with the leverage-share distribution last.
suppressPackageStartupMessages(library(patchwork))

make_trim_panel = function(oc, tag, show_y_labels) {
  d = leverage_trim_est %>%
    dplyr::filter(outcome == oc) %>%
    dplyr::mutate(sample = factor(sample, levels = trim_panel_levels))
  panel_title = if (identical(oc, "pm1mun")) {
    expression(paste("(g) ", PM[1]))
  } else {
    paste0(tag, " ", unname(main_outcome_labels[oc]))
  }
  ggplot(d, aes(y = sample, x = estimate, color = sample)) +
    geom_vline(
      xintercept = 0,
      color = "black",
      linewidth = 0.3,
      linetype = "dashed"
    ) +
    geom_segment(
      aes(x = conf.low95, xend = conf.high95, yend = sample),
      linewidth = 1.4,
      alpha = 0.28
    ) +
    geom_segment(
      aes(x = conf.low90, xend = conf.high90, yend = sample),
      linewidth = 1.4,
      alpha = 0.68
    ) +
    geom_point(size = 2.1) +
    scale_color_manual(values = trim_panel_cols, guide = "none") +
    scale_y_discrete(limits = rev(trim_panel_levels), drop = FALSE) +
    labs(
      title = panel_title,
      x = "Estimate",
      y = NULL
    ) +
    theme_minimal(base_size = 9) +
    theme(
      panel.grid.major.y = element_blank(),
      axis.text.y = if (show_y_labels) {
        element_text(color = "black", size = 7.5)
      } else {
        element_blank()
      },
      axis.ticks.y = element_blank(),
      plot.title = element_text(hjust = 0, size = 9.5),
      legend.position = "none"
    )
}

trim_coefficient_panels = Map(
  function(oc, tag, panel_index) {
    make_trim_panel(
      oc,
      tag,
      show_y_labels = panel_index %in% c(1, 4, 7)
    )
  },
  main_outcomes,
  paste0("(", letters[seq_along(main_outcomes)], ")"),
  seq_along(main_outcomes)
)

# leverage-share histogram,
p_i = fig_leverage_hist +
  labs(
    title = "(i) Leverage-share distribution",
    x = "Leverage share (% of total, log scale)",
    y = NULL
  ) +
  theme_minimal(base_size = 9) +
  theme(
    plot.title = element_text(hjust = 0, size = 9.5),
    legend.position = "none"
  )

fig_leverage_trim_panel = patchwork::wrap_plots(
  c(trim_coefficient_panels, list(p_i)),
  ncol = 3
)
print(fig_leverage_trim_panel)
```


### Leave-one-microregion-out

Each red point comes from re-estimating the headline 2SLS after deleting one microregion.


```{r RobustOneOut, include=FALSE}
# Leave-one-microregion-out robustness -------------------------------------
outcome_microreg = setNames(
  lapply(main_outcomes, function(oc) {
    dta_normal %>%
      dplyr::filter(reg_sample == 1, !is.na(.data[[oc]]), !is.na(microreg)) %>%
      dplyr::distinct(microreg) %>%
      dplyr::pull(microreg) %>%
      as.character() %>%
      sort()
  }),
  main_outcomes
)

loo_microregions = sort(unique(unlist(outcome_microreg)))

# Pass every dependency explicitly so future does not export the Rmd environment.
loo_worker = function(drop_microreg, data_worker, outcomes_worker,
                       outcome_microreg_worker, covariates_worker,
                       library_paths_worker) {
  .libPaths(library_paths_worker)
  fixest::setFixest_nthreads(1)
  data_worker = dplyr::filter(
    data_worker, reg_sample == 1, as.character(microreg) != drop_microreg
  )

  purrr::map_dfr(outcomes_worker, function(outcome_i) {
    if (!(drop_microreg %in% outcome_microreg_worker[[outcome_i]])) return(NULL)

    est = fixest::feols(
      stats::as.formula(paste0(
        outcome_i, " ~ ", paste(covariates_worker, collapse = " + "),
        " | codmun + interaction(uf, year)",
        " | starlink ~ coverarea:year"
      )),
      data = data_worker,
      cluster = ~microreg,
      notes = FALSE,
      nthreads = 1
    )
    tab = dplyr::filter(broom::tidy(est), term == "fit_starlink")
    ci95 = as.numeric(stats::confint(est, "fit_starlink", level = 0.95)[1, ])
    ci90 = as.numeric(stats::confint(est, "fit_starlink", level = 0.90)[1, ])

    tibble::tibble(
      outcome = outcome_i, estimate = tab$estimate,
      conf.low95 = ci95[1], conf.high95 = ci95[2],
      conf.low90 = ci90[1], conf.high90 = ci90[2],
      dropped_microreg = drop_microreg
    )
  })
}
environment(loo_worker) = globalenv()

loo_est = local({
  future::plan(
    future::multisession,
    workers = min(robust_max_workers, length(loo_microregions))
  )
  on.exit(future::plan(future::sequential), add = TRUE)
  dplyr::bind_rows(future_lapply(
    X = loo_microregions,
    FUN = loo_worker,
    data_worker = dta_normal,
    outcomes_worker = main_outcomes,
    outcome_microreg_worker = outcome_microreg,
    covariates_worker = model_controls,
    library_paths_worker = .libPaths(),
    future.globals = FALSE,
    future.seed = TRUE
  ))
})

# Order microregions by state and show state blocks below each panel.
uf_code_to_sigla = c(
    "11" = "RO",
    "12" = "AC",
    "13" = "AM",
    "14" = "RR",
    "15" = "PA",
    "16" = "AP",
    "17" = "TO",
    "21" = "MA",
    "51" = "MT"
)

make_loo_fig = function(oc, panel_index) {
    d = loo_est %>%
      dplyr::filter(outcome == oc, !is.na(estimate)) %>%
      dplyr::mutate(
        uf_code = substr(dropped_microreg, 1, 2),
        uf = unname(uf_code_to_sigla[uf_code])
      ) %>%
      dplyr::arrange(uf_code, dropped_microreg) %>%
      dplyr::mutate(
        x = dplyr::row_number(),
        uf = dplyr::coalesce(uf, uf_code)
      )
    panel_title = if (identical(oc, "pm1mun")) {
      bquote(paste("(", .(letters[panel_index]), ") ", PM[1]))
    } else {
      main_outcome_titles[[oc]]
    }

    # Draw a bracket for each state's block of microregions.
    uf_axis = d %>%
      dplyr::group_by(uf) %>%
      dplyr::summarise(start = min(x), end = max(x), .groups = "drop") %>%
      dplyr::arrange(start) %>%
      dplyr::mutate(mid = (start + end) / 2)
    uf_bound_df = tibble::tibble(
      b = sort(unique(c(uf_axis$start - 0.5, uf_axis$end + 0.5)))
    )

    # Place the brackets below the confidence intervals and zero.
    y_lo = min(c(d$conf.low95, 0), na.rm = TRUE)
    y_hi = max(c(d$conf.high95, 0), na.rm = TRUE)
    rng = y_hi - y_lo
    if (!is.finite(rng) || rng <= 0) {
      rng = 1
    }
    y_br = y_lo - 0.12 * rng # bracket baseline
    cap = 0.04 * rng # half-height of the end caps

    ggplot(d, aes(x = x, y = estimate)) +
      geom_hline(
        yintercept = 0,
        color = "black",
        linewidth = 0.4,
        linetype = "dashed"
      ) +
      # Draw the 95% interval behind the darker 90% interval.
      geom_linerange(
        aes(ymin = conf.low95, ymax = conf.high95),
        color = "grey80",
        linewidth = 1.2,
        na.rm = TRUE
      ) +
      geom_linerange(
        aes(ymin = conf.low90, ymax = conf.high90),
        color = "grey40",
        linewidth = 1.2,
        na.rm = TRUE
      ) +
      geom_point(color = "red", size = 0.8, na.rm = TRUE) +
      geom_segment(
        data = uf_axis,
        inherit.aes = FALSE,
        aes(x = start - 0.5, xend = end + 0.5, y = y_br, yend = y_br),
        linewidth = 0.4
      ) +
      geom_segment(
        data = uf_bound_df,
        inherit.aes = FALSE,
        aes(x = b, xend = b, y = y_br - cap, yend = y_br + cap),
        linewidth = 0.4
      ) +
      geom_text(
        data = uf_axis,
        inherit.aes = FALSE,
        aes(x = mid, y = y_br - 2.4 * cap, label = uf),
        size = 3,
        vjust = 1
      ) +
      scale_x_continuous(expand = ggplot2::expansion(mult = 0.01)) +
      scale_y_continuous(
        breaks = pretty(c(y_lo, y_hi), n = 5),
        expand = ggplot2::expansion(mult = c(0.16, 0.05))
      ) +
      coord_cartesian(clip = "off") +
      labs(
        title = panel_title,
        x = NULL,
        y = main_outcome_ylabs[[oc]]
      ) +
      theme_minimal() +
      theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.title = element_text(size = 9, hjust = 0),
        axis.title.y = element_text(size = 8),
        axis.text.y = element_text(size = 7),
        plot.margin = margin(4, 5, 6, 5),
        legend.position = "none",
        text = element_text(size = 8)
      )
}

loo_panels = Map(make_loo_fig, main_outcomes, seq_along(main_outcomes))
fig_lomro_8 = ggpubr::ggarrange(
  plotlist = loo_panels, ncol = 2, nrow = 4, align = "hv"
)
```

```{r LooFigInline, message=FALSE, warning=FALSE, fig.width=10, fig.height=8, echo=FALSE}
print(fig_lomro_8)
```



### Leave-one-state-out


- Drop each of the 9 Legal-Amazon states (UF) in turn and re-estimate.


```{r RobustOneStateOut, include=FALSE}
# Leave-one-STATE-out robustness ------------------------------------------
loso_states = sort(unique(dta_normal$uf[dta_normal$reg_sample == 1]))

# 9 states x 8 outcomes is light enough to run sequentially.
loso_est = purrr::map_dfr(loso_states, function(drop_uf) {
  data_i = dplyr::filter(dta_normal, uf != drop_uf)
  purrr::map_dfr(main_outcomes, function(outcome_i) {
    estimate_starlink(data_i, outcome_i) %>%
      dplyr::mutate(dropped_uf = drop_uf)
  })
})

uf_levels = unname(uf_code_to_sigla)
loso_plot_df = bind_rows(
  robust_full %>%
    transmute(outcome, estimate, conf.low95, conf.high95,
              conf.low90, conf.high90, uf = "Full"),
  loso_est %>%
    transmute(
      outcome, estimate, conf.low95, conf.high95, conf.low90, conf.high90,
      uf_code = as.character(dropped_uf),
      uf = dplyr::coalesce(unname(uf_code_to_sigla[uf_code]), uf_code)
    )
) %>%
  dplyr::mutate(uf = factor(uf, levels = c("Full", uf_levels)))

# The number below each state abbreviation is the number of municipalities that
# state drop excludes from the outcome's own estimation sample. The number below
# Full is that outcome's full-sample municipality count. These labels must be
# outcome-specific because the UPF-overlap and Spatial-HHI samples are smaller.
loso_state_counts = purrr::map_dfr(main_outcomes, function(oc) {
  dplyr::filter(
    dta_normal,
    reg_sample == 1,
    !is.na(.data[[oc]])
  ) %>%
    dplyr::distinct(codmun, uf) %>%
    dplyr::mutate(
      uf_code = as.character(uf),
      uf_sigla = dplyr::coalesce(
        unname(uf_code_to_sigla[uf_code]),
        uf_code
      )
    ) %>%
    dplyr::count(uf_sigla, name = "n_mun") %>%
    dplyr::mutate(outcome = oc)
})

loso_x_labels = setNames(
  lapply(main_outcomes, function(oc) {
    counts = dplyr::filter(loso_state_counts, outcome == oc)
    n_by_uf = setNames(counts$n_mun, counts$uf_sigla)[uf_levels]
    n_by_uf[is.na(n_by_uf)] = 0
    c(
      Full = paste0("Full\n(", sum(n_by_uf), ")"),
      setNames(paste0(uf_levels, "\n(", n_by_uf, ")"), uf_levels)
    )
  }),
  main_outcomes
)

make_loso_fig = function(oc, panel_index) {
  d = dplyr::filter(loso_plot_df, outcome == oc)
  panel_title = if (identical(oc, "pm1mun")) {
    bquote(paste("(", .(letters[panel_index]), ") ", PM[1]))
  } else {
    main_outcome_titles[[oc]]
  }
  ggplot(d, aes(x = uf, y = estimate)) +
    geom_hline(
      yintercept = 0,
      color = "black",
      linewidth = 0.4,
      linetype = "dashed"
    ) +
    # 95% outer whisker (light grey) with 90% inner overlay (darker grey),
    # matching the placebo / trim figures' grey-and-red palette.
    geom_linerange(
      aes(ymin = conf.low95, ymax = conf.high95),
      color = "grey80",
      linewidth = 2,
      na.rm = TRUE
    ) +
    geom_linerange(
      aes(ymin = conf.low90, ymax = conf.high90),
      color = "grey40",
      linewidth = 2,
      na.rm = TRUE
    ) +
    geom_point(color = "red", size = 1.5, na.rm = TRUE) +
    scale_x_discrete(drop = FALSE, labels = loso_x_labels[[oc]]) +
    labs(
      title = panel_title,
      x = NULL,
      y = main_outcome_ylabs[[oc]]
    ) +
    theme_minimal() +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      plot.title = element_text(size = 12),
      plot.margin = margin(6, 10, 8, 10),
      legend.position = "none",
      text = element_text(size = 10)
    )
}

loso_panels = Map(make_loso_fig, main_outcomes, seq_along(main_outcomes))
fig_loso_4 = ggpubr::ggarrange(
  plotlist = loso_panels, ncol = 2, nrow = 4, align = "hv"
)
```

```{r LosoFigInline, message=FALSE, warning=FALSE, fig.width=10, fig.height=8, echo=FALSE}
print(fig_loso_4)
```



## Heterogeneity

```{r Heterogeneity, message=FALSE, warning=FALSE, R.options = list(width = 1200), echo=FALSE, fig.height=9, fig.width=10}
# Both reported heterogeneity variables are binary indicators.
groups = c("faction", "uc_dummy")
levels_vec = c(1, 0)
labels_vec = c("Yes", "No")

for (group in groups) {
  # Yes-No difference test from a single fully interacted IV model.
  saturated_iv_p = function(y) {
    d = dta %>%
      dplyr::filter(reg_sample == 1, .data[[group]] %in% levels_vec) %>%
      dplyr::mutate(grp = as.numeric(.data[[group]]))
    sat_covs = c(model_controls, paste0(model_controls, ":grp"))
    f = as.formula(paste0(
      y,
      " ~ ",
      paste(sat_covs, collapse = " + "),
      " | interaction(codmun, grp) + interaction(uf, year, grp)",
      " | starlink + starlink:grp ~ coverarea:year + coverarea:year:grp"
    ))
    est = feols(f, data = d, cluster = ~microreg, notes = FALSE)
    unname(est$coeftable["fit_starlink:grp", 4])
  }

  # Four columns: outcome, Yes, No, and their difference.
  n_rows = 3 + 2 * length(main_outcomes) + 2
  result = data.frame(matrix(ncol = 4, nrow = n_rows, data = ""))
  result[1, 2:3] = labels_vec
  result[2, 2:3] = "(II)"
  result[2, 4] = "Diff."
  result[3, 1] = "Outcomes"

  i = 4
  for (vardep in main_outcomes) {
    rownames(result)[i] = paste0(vardep, "_est")
    rownames(result)[i + 1] = paste0(vardep, "_se")
    result[i, 1] = heterogeneity_label(vardep)
    i = i + 2
  }

  # Closing rule under the last outcome; the N row is filled in below.
  result[4 + 2 * length(main_outcomes), ] = "-------"

  for (y in main_outcomes) {
    model = as.formula(paste0(
      y, " ~ ", paste(model_controls, collapse = " + "),
      " | codmun + interaction(uf, year)",
      " | starlink ~ coverarea:year"
    ))
    est_betas = numeric(length(levels_vec))

    for (lv in seq_along(levels_vec)) {
      est = feols(
        model,
        data = dplyr::filter(
          dta, reg_sample == 1, .data[[group]] == levels_vec[lv]
        ),
        cluster = ~microreg,
        notes = FALSE
      )
      tab = dplyr::filter(broom::tidy(est), term == "fit_starlink")
      result[paste0(y, "_est"), 1 + lv] = paste0(
        sprintf("%.3f", tab$estimate),
        case_when(
          tab$p.value <= .01 ~ "***",
          tab$p.value <= .05 ~ "**",
          tab$p.value <= .10 ~ "*",
          TRUE ~ ""
        )
      )
      result[paste0(y, "_se"), 1 + lv] =
        paste0("(", sprintf("%.3f", tab$std.error), ")")
      est_betas[lv] = tab$estimate
    }

    p_diff = saturated_iv_p(y)
    result[paste0(y, "_est"), 4] = paste0(
      sprintf("%.3f", est_betas[1] - est_betas[2]),
      case_when(
        p_diff <= .01 ~ "***",
        p_diff <= .05 ~ "**",
        p_diff <= .10 ~ "*",
        TRUE ~ ""
      )
    )
  }

  rownames(result) = 1:nrow(result)
  colnames(result) = NULL

  # N row
  n_row = nrow(result)
  result[n_row - 1, 1] = "-------"
  result[n_row, 1] = "N"
  for (lv in seq_along(levels_vec)) {
    n_obs = dta %>%
      filter(reg_sample == 1 & .data[[group]] == levels_vec[lv]) %>%
      nrow()
    result[n_row, 1 + lv] = format(n_obs, big.mark = ",")
  }

  print(paste(rep("=", 40), collapse = " "))
  print(paste0(
    "## Heterogeneity by ",
    var_names$varname[match(group, var_names$var)]
  ))
  print(format(result, justify = "left"))

}
```



---

# PRODES Year

This section produces the manuscript appendix results through official PRODES
2025 (August 2024--July 2025).

```{r DataNormalizationProdes, echo=FALSE}
# The PRODES-year panel uses July SPEI-12 and omits 2023 from the IV basis.
model_controls = sub("spei12_dec", "spei12_jul", model_controls, fixed = TRUE)
covlist = list("1", model_controls)
prodes_iv_names = paste0("caP_", 2024:2025)
prodes_iv_rhs = paste(prodes_iv_names, collapse = " + ")

dta = readRDS(paste0(DATA_DIR, "dataset_prodesyr.RDS")) %>%
  filter(year >= 2017 & year <= 2025 & !is.na(coverarea)) %>%
  mutate(
    # Restore the FGGD pasture-suitability index to its original 0-100 scale
    pasture_suit_mun = pasture_suit_mun / 100,
    year = factor(year, levels = as.character(2025:2017)),
    way_rail_oper = way_rail,
    caP_2024 = coverarea * (as.integer(as.character(year)) == 2024),
    caP_2025 = coverarea * (as.integer(as.character(year)) == 2025)
  )

# Outcome normalization follows the calendar-year panel
area_list = colnames(dta)[startsWith(colnames(dta), "area") | startsWith(colnames(dta), "way")]
for (y in area_list) {
  dta[[paste0("rm_", y)]] = (dta[[y]] / dta[["areamun"]]) * 100
  dta[[paste0("r_", y)]] = if (startsWith(y, "way")) {
    dta[[paste0("rm_", y)]]
  } else {
    (dta[[y]] / dta$area_forest) * 100
  }
}

n_list = colnames(dta)[startsWith(colnames(dta), "n_")]
for (y in n_list) {
  dta[[paste0("rm_", y)]] = (dta[[y]] / dta[["areamun"]]) * 100
  dta[[paste0("r_", y)]] = (dta[[y]] / dta$area_forest) * 100
}

dta = dta %>%
  mutate(
    areamun = areamun / 1000,
    area_forest = area_forest / 1000,
    dist_coast = dist_coast / 1000,
    dist_brasilia = dist_brasilia / 1000
  ) %>%
  select(-any_of(c("r_areamun", "rm_areamun", "r_area_forest")))

# Native municipal PRODES outcomes from polygons
prodes_varlist = c(
  "r_n_mun_prodes_defo",
  "r_n_mun_prodes_defo_ccut",
  "r_n_mun_prodes_defo_cces",
  "r_n_mun_prodes_defo_ccwv",
  "r_n_mun_prodes_defo_dpdp",
  "r_n_mun_prodes_defo_mine",
  "r_area_mun_prodes_defo",
  "r_area_mun_prodes_defo_ccut",
  "r_area_mun_prodes_defo_cces",
  "r_area_mun_prodes_defo_ccwv",
  "r_area_mun_prodes_defo_dpdp",
  "r_area_mun_prodes_defo_mine"
)

# Olea-Pflueger effective F for each outcome's estimation sample.
op_F_prodes_outcome = function(y) {
  d = dplyr::filter(dta, reg_sample == 1, !is.na(.data[[y]]))
  controls = split_controls_for_ivdiag(model_controls, d)
  op_F_clustered(
    data = d,
    yvar = y,
    formula_controls = controls$formula,
    bare_controls = controls$bare,
    iv = prodes_iv_names
  )
}
```

## Estimations

- Model `r as.character(as.roman(1))`: no covariates
- Model `r as.character(as.roman(length(covlist)))`: all covariates

```{r EstimationsProdes, message=FALSE, warning=FALSE, R.options = list(width = 1200), echo=FALSE}
# Estimate the 12 native PRODES outcomes on the August-July reference year.
result = data.frame(matrix(
  ncol = 1 + length(covlist),
  nrow = 2 + 2 * length(prodes_varlist) + 1,
  data = ""
))
result[1, 1 + ceiling(length(covlist) / 2)] = "Starlink"
result[2, 1] = "Outcomes"
result[2, 2:ncol(result)] = paste0("  (", as.roman(seq_along(covlist)), ")")
result = cbind(result, data.frame(ci95 = "", t = "", p = "", N = "", opF = ""))
result[2, (ncol(result) - 4):ncol(result)] = c("95% CI", "t-stat", "p-val", "N", "OP F")
result[3 + 2 * length(prodes_varlist), ] = "-------"

i = 3
for (vardep in prodes_varlist) {
  rownames(result)[i] = paste0(vardep, "_est")
  rownames(result)[i + 1] = paste0(vardep, "_se")
  result[i, 1] = var_names$varname[match(vardep, var_names$var)]
  i = i + 2
}

for (m in seq_along(covlist)) {
  for (y in prodes_varlist) {
    tab = estimate_starlink(
      dta, y, iv_rhs = prodes_iv_rhs, controls = covlist[[m]]
    )
    result[paste0(y, "_est"), m + 1] = paste0(
      sprintf("%.3f", tab$estimate),
      case_when(
        tab$p.value <= .01 ~ "***",
        tab$p.value <= .05 ~ "**",
        tab$p.value <= .10 ~ "*",
        TRUE ~ ""
      )
    )
    result[paste0(y, "_se"), m + 1] = paste0(
      "(", sprintf("%.3f", tab$std.error), ")"
    )

    if (m == length(covlist)) {
      result[paste0(y, "_est"), "ci95"] = paste0(
        "[", sprintf("%.3f", tab$conf.low95), ", ",
        sprintf("%.3f", tab$conf.high95), "]"
      )
      result[paste0(y, "_est"), "t"] = sprintf("%.3f", tab$statistic)
      result[paste0(y, "_est"), "p"] = sprintf("%.3f", tab$p.value)
      result[paste0(y, "_est"), "N"] = formatC(tab$N, format = "d", big.mark = ",")
      result[paste0(y, "_est"), "opF"] = sprintf("%.1f", op_F_prodes_outcome(y))
    }
  }
}
rownames(result) = 1:nrow(result)
colnames(result) = NULL

print(format(
  result[, -c((ncol(result) - 4):(ncol(result) - 3))],
  justify = "left"
))
```

## Session information

```{r SessionInfo, echo=FALSE}
sessionInfo()
```
