---
title: "Building the analytical dataset for 'The Effects of Starlink Adoption on Forest Degradation in the Amazon'"
author: "Anonymized for peer review"
date: "September, 2026"
output: html_document
---

## SETUP
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  echo = TRUE
)

# Clear the environment before construction.
rm(list = ls())
gc()
```


## 0. INITIALIZATION
```{r 0, include=FALSE}
max_workers = max(1L, min(9L, parallel::detectCores(logical = FALSE) - 1L))

DETER_zip = "INPE/deter-amz-public-2025set01.zip"

# Data locations relative to this script.
DATA_DIR = "../data/"
DV = paste0(DATA_DIR, "processed/")
DRAW = paste0(DATA_DIR, "raw/")
DS = paste0(DV, "tenures/")

dir.create(DS, recursive = TRUE, showWarnings = FALSE)


library(tidyverse)
library(sf)
library(readxl)
library(stringi)
library(exactextractr)
library(terra)
library(future.apply)

sum_coverage = function(x) {
  list(
    x %>%
      filter(is.finite(value), is.finite(coverage_area)) %>%
      mutate(
        total_area = sum(coverage_area, na.rm = TRUE),
        proportion = coverage_area / total_area,
        product = value * proportion
      ) %>%
      select(product)
  )
}

# Official CRS for computing areas (SIRGAS 2000 - conic Albers equiv, in meters)
# https://biblioteca.ibge.gov.br/visualizacao/livros/liv102169.pdf (page 21)
proj_wk = 'PROJCS["Conica_Equivalente_de_Albers_Brasil",  
    GEOGCS["GCS_SIRGAS2000",  
      DATUM["D_SIRGAS2000",  
        SPHEROID["Geodetic_Reference_System_of_1980",6378137,298.2572221009113]],  
      PRIMEM["Greenwich",0],  
      UNIT["Degree",0.017453292519943295]],  
    PROJECTION["Albers"],  
    PARAMETER["standard_parallel_1",-2],  
    PARAMETER["standard_parallel_2",-22],  
    PARAMETER["latitude_of_origin",-12],  
    PARAMETER["central_meridian",-54],  
    PARAMETER["false_easting",5000000],  
    PARAMETER["false_northing",10000000],  
    UNIT["Meter",1]]'

# Mojui dos Campos is merged into Santarem throughout the panel.
MOJUI_CODMUN = 1504752
SANTAREM_CODMUN = 1506807

merge_codmun = function(x) ifelse(x == MOJUI_CODMUN, SANTAREM_CODMUN, x)



# Structural exclusion applied before the tenure masks are built.
SICAR_MAX_HA = 1e5  # a registration larger than this is not a property

ufs_sicar = c("AC", "AM", "AP", "MA", "MT", "PA", "RO", "RR", "TO")

# SICAR geometry helpers normalize empty overlay results to one MULTIPOLYGON.
EMPTY_MP = st_sfc(st_multipolygon(), crs = proj_wk)

is_nothing = function(g) { length(g) == 0 || all(st_is_empty(g))}

# Retain only polygonal parts produced by spatial overlays.
as_mp = function(g) {
  if (is.null(g) || is_nothing(g)) return(EMPTY_MP)
  if (any(st_geometry_type(g) == "GEOMETRYCOLLECTION")) {
    g = st_collection_extract(g, "POLYGON")
  }
  g = g[st_geometry_type(g) %in% c("POLYGON", "MULTIPOLYGON")]
  if (is_nothing(g)) return(EMPTY_MP)
  out = tryCatch(st_make_valid(st_union(g)), error = function(e) {
    st_make_valid(st_union(st_make_valid(lwgeom::st_snap_to_grid(g, 0.01))))
  })
  if (is_nothing(out)) EMPTY_MP else st_cast(out, "MULTIPOLYGON", warn = FALSE)
}

u_or_empty = function(...) {
  as_mp(do.call(c, Filter(Negate(is.null), list(...))))
}

# Keep the original geometry and flag the municipality if both differences fail.
safe_diff = function(x, y) {
  if (is_nothing(x) || is_nothing(y)) {
    return(list(g = if (is_nothing(x)) EMPTY_MP else x, failed = FALSE))
  }
  failed = FALSE
  out = tryCatch(
    st_make_valid(st_difference(x, y)),
    error = function(e) tryCatch(
      st_make_valid(st_difference(
        st_make_valid(lwgeom::st_snap_to_grid(x, 0.01)),
        st_make_valid(lwgeom::st_snap_to_grid(y, 0.01))
      )),
      error = function(e2) {
        failed <<- TRUE
        x
      }
    )
  )
  list(g = as_mp(out), failed = failed)
}

# Intersection kept only where it is a real piece of land: everything under
# 1 ha is dropped as boundary noise between two independently digitized layers.
safe_over = function(x, y, sliver_m2 = 1e4) {
  if (is_nothing(x) || is_nothing(y)) return(EMPTY_MP)
  o = tryCatch(st_make_valid(st_intersection(x, y)), error = function(e) {
    tryCatch(st_make_valid(st_intersection(
      st_make_valid(lwgeom::st_snap_to_grid(x, 0.01)),
      st_make_valid(lwgeom::st_snap_to_grid(y, 0.01))
    )), error = function(e2) EMPTY_MP)
  })
  o = as_mp(o)                     # one clean MULTIPOLYGON, or nothing
  if (is_nothing(o)) return(EMPTY_MP)
  parts = tryCatch(st_cast(o, "POLYGON", warn = FALSE),
                   error = function(e) EMPTY_MP)
  if (is_nothing(parts)) return(EMPTY_MP)
  a = as.numeric(st_area(parts))
  keep = a >= sliver_m2
  if (any(keep)) as_mp(parts[keep]) else EMPTY_MP
}

km2 = function(g) if (is_nothing(g)) 0 else sum(as.numeric(st_area(g))) / 1e6

# Reduce each feature to one polygonal geometry and drop arealess results.
keep_polygons = function(x) {
  if (nrow(x) == 0) return(x)
  g = st_geometry(x)
  is_gc = st_geometry_type(g) == "GEOMETRYCOLLECTION"
  if (any(is_gc)) {
    crs_x = st_crs(g)
    g = st_sfc(lapply(seq_along(g), function(k) {
      if (!is_gc[k]) return(g[[k]])
      part = suppressWarnings(
        st_collection_extract(st_sfc(g[[k]], crs = crs_x), "POLYGON")
      )
      if (length(part) == 0 || all(st_is_empty(part))) {
        st_multipolygon()
      } else {
        st_combine(part)[[1]]
      }
    }), crs = crs_x)
    st_geometry(x) = g
  }
  x = x[st_geometry_type(st_geometry(x)) %in% c("POLYGON", "MULTIPOLYGON"), ]
  if (nrow(x) == 0) return(x)
  x[!st_is_empty(st_geometry(x)) &
      as.numeric(st_area(st_geometry(x))) > 0, ]
}


# Build the tenure layers from SICAR and public polygons.

classify_sicar_stock = function(x) {
  module_ref = x %>%
    st_drop_geometry() %>%
    filter(!is.na(mod_fiscal), mod_fiscal > 0,
           !is.na(num_area), num_area > 0) %>%
    mutate(module_ha = num_area / mod_fiscal)

  module_by_mun = module_ref %>%
    group_by(uf, municipio) %>%
    summarise(module_ha = median(module_ha), .groups = "drop")

  module_by_uf = module_ref %>%
    group_by(uf) %>%
    summarise(module_uf = median(module_ha), .groups = "drop")

  x %>%
    left_join(module_by_mun, by = c("uf", "municipio")) %>%
    left_join(module_by_uf, by = "uf") %>%
    mutate(
      mf_used = ifelse(
        !is.na(mod_fiscal) & mod_fiscal > 0,
        mod_fiscal,
        num_area / coalesce(module_ha, module_uf)
      ),
      size_class = case_when(
        is.na(mf_used) | !is.finite(mf_used) | mf_used <= 0 ~ "unclassified",
        mf_used > 4 ~ "large",
        TRUE ~ "small"
      )
    ) %>%
    select(-module_ha, -module_uf)
}

build_sicar_stock_uf = function(uf, mun_zone, sicar_dir, sicar_work_dir) {
  zip_path = file.path(sicar_dir, paste0("AREA_IMOVEL_", uf, ".zip"))
  d = st_read(paste0("/vsizip/", zip_path, "/AREA_IMOVEL_1.shp"), quiet = TRUE)

  # Retain private rural registrations present by 2021.
  d = d %>%
    mutate(
      uf = uf,
      cria = lubridate::dmy(dat_criaca, quiet = TRUE),
      mod_fiscal = suppressWarnings(as.numeric(mod_fiscal)),
      num_area = suppressWarnings(as.numeric(num_area))
    ) %>%
    filter(
      ind_tipo == "IRU",
      !is.na(cria),
      lubridate::year(cria) <= 2021
    )

  d = d %>%
    st_make_valid() %>%
    st_transform(crs = proj_wk) %>%
    lwgeom::st_snap_to_grid(0.01) %>%
    st_make_valid() %>%
    keep_polygons() %>%
    st_cast("MULTIPOLYGON", warn = FALSE) %>%
    mutate(
      area_m2 = as.numeric(st_area(st_geometry(.))),
      geom_ha = area_m2 / 1e4
    )

  d = d %>%
    filter(geom_ha <= SICAR_MAX_HA) %>%
    classify_sicar_stock() %>%
    filter(ind_status != "CA")

  # Drop sub-square-meter artifacts with no fiscal-module classification.
  d = d %>%
    filter(!(size_class == "unclassified" & area_m2 < 1))
  if (any(d$size_class == "unclassified")) {
    stop(uf, ": fiscal-module size is missing for ",
         sum(d$size_class == "unclassified"), " retained registrations")
  }

  d %>%
    select(cod_imovel, size_class, mf_used, area_m2) %>%
    st_intersection(mun_zone) %>%
    st_make_valid() %>%
    keep_polygons() %>%
    saveRDS(file.path(sicar_work_dir, paste0("stock_pieces_", uf, ".RDS")))
}

# Municipal polygons are the common clipping frame for every tenure layer.
# The Legal Amazon source can contain Santarem and Mojui separately; the
# analysis uses one merged unit throughout.
pop_2022 = read.csv(
  paste0(DRAW, "IBGE/tabela4714.csv"),
  sep = ",", quote = "\"", skip = 3
)
names(pop_2022)[1] = "codmun"
pop_2022 = pop_2022 %>%
  filter(grepl("^[0-9]{7}$", codmun)) %>%
  transmute(
    codmun = as.numeric(codmun),
    pop = as.numeric(X2022)
  )

amz = st_read(
  paste0("/vsizip/", DRAW, "IBGE/municipalities_legal_amazon.zip/municipalities_legal_amazon.shp"),
  quiet = TRUE
) %>%
  st_make_valid() %>%
  st_transform(crs = proj_wk) %>%
  lwgeom::st_snap_to_grid(0.01) %>%
  st_make_valid() %>%
  mutate(
    areamun = as.numeric(units::set_units(st_area(.), km^2)),
    codmun = as.numeric(geocodigo),
    uf = as.numeric(substr(geocodigo, 1, 2))
  ) %>%
  left_join(pop_2022, by = "codmun") %>%
  arrange(codmun) %>%
  mutate(
    codmun = merge_codmun(codmun),
    namemun = ifelse(codmun == SANTAREM_CODMUN, "Santarem + Mojui dos Campos", nome)
  ) %>%
  group_by(codmun, namemun, uf) %>%
  summarise(
    areamun = sum(areamun),
    pop = sum(pop),
    .groups = "drop"
  ) %>%
  st_make_valid() %>%
  arrange(codmun)

uc = st_read(
  paste0("/vsizip/", DRAW, "ICMBio/conservation_units_legal_amazon.zip/",
         "conservation_units_legal_amazon.shp"),
  quiet = TRUE
) %>%
  st_make_valid() %>%
  st_transform(crs = proj_wk) %>%
  lwgeom::st_snap_to_grid(0.01) %>%
  st_make_valid() %>%
  select(id_uc = id, sphere_uc = esfera, class_uc = categoria, year = ano_cria) %>%
  st_intersection(amz %>% select(codmun)) %>%
  st_make_valid() %>%
  keep_polygons() %>%
  mutate(
    areauc = as.numeric(units::set_units(st_area(.), km^2)),
    year = suppressWarnings(as.numeric(
      stringr::str_extract(as.character(year), "[0-9]{4}")
    ))
  ) %>%
  select(codmun, id_uc, year, class_uc, areauc) %>%
  arrange(codmun)

ti = st_read(
  paste0("/vsizip/", DRAW, "FUNAI/indigenous_area_legal_amazon.zip/indigenous_area_legal_amazon.shp"),
  quiet = TRUE
) %>%
  st_make_valid() %>%
  st_transform(crs = proj_wk) %>%
  lwgeom::st_snap_to_grid(0.01) %>%
  st_make_valid() %>%
  select(id_ti = terrai_cod) %>%
  st_intersection(amz %>% select(codmun)) %>%
  st_make_valid() %>%
  keep_polygons() %>%
  mutate(areati = as.numeric(units::set_units(st_area(.), km^2))) %>%
  select(codmun, id_ti, areati) %>%
  arrange(codmun)

sett = st_read(
  paste0("/vsizip/", DRAW, "INCRA/Assentamento Brasil.zip/Assentamento Brasil.shp"),
  quiet = TRUE
) %>%
  st_make_valid() %>%
  st_transform(crs = proj_wk) %>%
  lwgeom::st_snap_to_grid(0.01) %>%
  st_make_valid() %>%
  select(id_sett = cd_sipra, data_de_cr, type = forma_obte) %>%
  mutate(
    year = lubridate::year(lubridate::dmy(data_de_cr, quiet = TRUE)),
    month = lubridate::month(lubridate::dmy(data_de_cr, quiet = TRUE))
  ) %>%
  st_intersection(amz %>% select(codmun)) %>%
  st_make_valid() %>%
  keep_polygons() %>%
  mutate(areasett = as.numeric(units::set_units(st_area(.), km^2))) %>%
  select(codmun, id_sett, areasett) %>%
  arrange(codmun)

ndpf = st_read(
  paste0("/vsizip/", DRAW, "CNFP/CNFP_2020.zip/CNFP_2020.shp"),
  quiet = TRUE
) %>%
  filter(tipo == "TIPO B") %>%
  st_make_valid() %>%
  st_transform(crs = proj_wk) %>%
  lwgeom::st_snap_to_grid(0.01) %>%
  st_make_valid() %>%
  select(id_ndpf = OBJECTID, sphere = governo, type = protecao) %>%
  st_intersection(amz %>% select(codmun)) %>%
  st_make_valid() %>%
  keep_polygons() %>%
  mutate(areandpf = as.numeric(units::set_units(st_area(.), km^2))) %>%
  select(codmun, id_ndpf, sphere, type, areandpf) %>%
  arrange(codmun)

base_zones = list(
  mun = list(zone = amz, prefix = "mun"),
  ti = list(zone = ti, prefix = "ti"),
  uc = list(zone = uc, prefix = "uc"),
  sett = list(zone = sett, prefix = "sett"),
  ndpf = list(zone = ndpf, prefix = "ndpf")
)

mun_zone = base_zones$mun$zone %>% select(codmun)
saveRDS(mun_zone, paste0(DV, "mun_zone.RDS"))

sicar_dir = normalizePath(file.path(DRAW, "SICAR"), winslash = "/", mustWork = TRUE)
sicar_work_dir = normalizePath(DS, winslash = "/", mustWork = TRUE)

public_layers = list(
  ndpf = base_zones$ndpf$zone %>% select(codmun),
  ti = base_zones$ti$zone %>% select(codmun),
  uc = base_zones$uc$zone %>% select(codmun),
  sett = base_zones$sett$zone %>% select(codmun)
)
saveRDS(public_layers, file.path(sicar_work_dir, "public_layers.RDS"))

options(future.globals.maxSize = 8 * 1024^3)
future::plan(
  future::multisession,
  workers = min(max_workers, length(ufs_sicar))
)

invisible(future.apply::future_lapply(
  ufs_sicar,
  function(uf) build_sicar_stock_uf(uf, mun_zone, sicar_dir, sicar_work_dir),
  future.seed = TRUE,
  future.packages = c("sf", "dplyr", "lubridate", "lwgeom")
))


# State files are regrouped before unioning because a registration can cross
# a state border. Each municipality is assigned whole to one balanced batch.
all_pieces = bind_rows(lapply(ufs_sicar, function(uf) {
  readRDS(file.path(sicar_work_dir, paste0("stock_pieces_", uf, ".RDS")))
}))
N_SICAR_BATCHES = 18
batch_lookup = all_pieces %>%
  st_drop_geometry() %>%
  count(codmun, name = "n_pieces") %>%
  arrange(desc(n_pieces)) %>%
  mutate(batch = rep_len(seq_len(N_SICAR_BATCHES), n()))

all_pieces = all_pieces %>%
  left_join(batch_lookup %>% select(codmun, batch), by = "codmun")
for (batch in seq_len(N_SICAR_BATCHES)) {
  saveRDS(
    all_pieces %>% filter(batch == !!batch) %>% select(-batch),
    file.path(sicar_work_dir, paste0("stock_batch_", batch, ".RDS"))
  )
}
rm(all_pieces)
gc()

build_sicar_masks_batch = function(batch, sicar_work_dir) {
  pieces = readRDS(file.path(sicar_work_dir, paste0("stock_batch_", batch, ".RDS")))
  public = readRDS(file.path(sicar_work_dir, "public_layers.RDS"))

  lapply(sort(unique(pieces$codmun)), function(codmun_i) {
    p = pieces[pieces$codmun == codmun_i, ]
    claims_small = u_or_empty(st_geometry(p)[p$size_class == "small"])
    claims_large = u_or_empty(st_geometry(p)[p$size_class == "large"])

    public_mun = lapply(public, function(layer) {
      u_or_empty(st_geometry(layer)[layer$codmun == codmun_i])
    })
    any_public = u_or_empty(
      public_mun$ndpf, public_mun$ti, public_mun$uc, public_mun$sett
    )

    net_small = safe_diff(claims_small, any_public)
    net_large = safe_diff(claims_large, any_public)

    # `safe_over()` dissolves each intersection to one MULTIPOLYGON, so a DETER
    # feature can be counted at most once within each reported size category.
    overlap_ndpf_large = safe_over(claims_large, public_mun$ndpf)
    overlap_ndpf_small = safe_over(claims_small, public_mun$ndpf)

    list(
      codmun = codmun_i,
      geoms = list(
        sicar21_small = net_small$g,
        sicar21_large = net_large$g,
        over_ndpf_large = overlap_ndpf_large,
        over_ndpf_small = overlap_ndpf_small
      ),
      area = tibble::tibble(
        codmun = codmun_i,
        areasicar21_small = km2(net_small$g),
        areasicar21_large = km2(net_large$g),
        areaover_ndpf_large = km2(overlap_ndpf_large),
        areaover_ndpf_small = km2(overlap_ndpf_small)
      ),
      net_failed = net_small$failed || net_large$failed
    )
  })
}

future::plan(future::multisession, workers = min(max_workers, N_SICAR_BATCHES))
masks = future.apply::future_lapply(
  seq_len(N_SICAR_BATCHES),
  function(batch) build_sicar_masks_batch(batch, sicar_work_dir),
  future.seed = TRUE,
  future.packages = c("sf", "dplyr", "lwgeom")
) %>%
  unlist(recursive = FALSE)
future::plan(future::sequential)

if (any(vapply(masks, `[[`, logical(1), "net_failed"))) {
  warning("SICAR netting failed in at least one municipality; inspect masks")
}

mask_names = c(
  "sicar21_small", "sicar21_large",
  "over_ndpf_large", "over_ndpf_small"
)
mask_sf = lapply(setNames(mask_names, mask_names), function(mask_name) {
  keep = Filter(function(x) !is_nothing(x$geoms[[mask_name]]), masks)
  if (length(keep) == 0) return(NULL)
  out = st_sf(
    codmun = vapply(keep, `[[`, numeric(1), "codmun"),
    geometry = do.call(c, lapply(keep, function(x) x$geoms[[mask_name]])),
    crs = proj_wk
  )
  out
})

sicar_area = mun_zone %>%
  st_drop_geometry() %>%
  left_join(bind_rows(lapply(masks, `[[`, "area")), by = "codmun") %>%
  mutate(across(-codmun, ~ replace_na(.x, 0))) %>%
  arrange(codmun)

zones = c(
  unname(base_zones),
  lapply(mask_names, function(mask_name) {
    list(zone = mask_sf[[mask_name]], prefix = mask_name)
  })
) %>%
  Filter(function(z) !is.null(z$zone), .)

zones_prefixes = sapply(zones, `[[`, "prefix")

rm(
  pop_2022, amz, ti, uc, sett, ndpf, base_zones, public_layers,
  batch_lookup, masks, mask_sf, sicar_dir, sicar_work_dir
)
gc()


# ---- population within the Legal Amazon portion of each municipality ---------
# The Legal Amazon boundary cuts 21 Maranhao municipalities: for those the
# municipal polygon carries only part of the territory the 2022 census counted,
# so the full census population would overstate density.

mun_pos = which(zones_prefixes == "mun")
mun_poly = zones[[mun_pos]]$zone
municipal_population = mun_poly %>%
  st_drop_geometry() %>%
  select(codmun, pop)

partial_population_path = paste0(DV, "pop_clipped_21.RDS")

if (!file.exists(partial_population_path)) {
  CENSUS_DIR = paste0(DRAW, "IBGE/")
  MA_MESH = paste0(CENSUS_DIR, "MA_setores_CD2022.gpkg")
  AGG_ZIP = paste0(CENSUS_DIR, "Agregados_por_setores_basico_BR_20260520.zip")

  fetch_once = function(url, dest, min_bytes = 1e5) {
    if (file.exists(dest) && file.size(dest) > min_bytes) return(invisible(TRUE))
    cat("fetching", basename(dest), "...\n")
    download.file(url, dest, mode = "wb", quiet = TRUE, method = "libcurl")
  }
  fetch_once(
    paste0("https://geoftp.ibge.gov.br/organizacao_do_territorio/",
           "malhas_territoriais/malhas_de_setores_censitarios__divisoes_",
           "intramunicipais/censo_2022/setores/gpkg/UF/MA/MA_setores_CD2022.gpkg"),
    MA_MESH
  )

  # Definitive aggregates (2026-05-20), NOT the 2024-03 preliminaries.
  fetch_once(
    paste0("https://ftp.ibge.gov.br/Censos/Censo_Demografico_2022/",
           "Agregados_por_Setores_Censitarios/Agregados_por_Setor_csv/",
           "Agregados_por_setores_basico_BR_20260520.zip"),
    AGG_ZIP
  )

  # Tract population (v0001), read straight from the zip -- nothing extracted.
  # Maranhao only; a tract's first two digits are its UF.
  csv_in_zip = utils::unzip(AGG_ZIP, list = TRUE)$Name
  csv_in_zip = csv_in_zip[grepl("[.]csv$", csv_in_zip, ignore.case = TRUE)][1]
  setor_pop = readr::read_delim(
    unz(AGG_ZIP, csv_in_zip), delim = ";", quote = "\"",
    locale = readr::locale(encoding = "latin1", decimal_mark = ","),
    col_types = readr::cols(.default = readr::col_character()),
    progress = FALSE) %>%
    transmute(CD_SETOR = as.character(CD_SETOR),
              pop_setor = suppressWarnings(as.numeric(v0001))) %>%
    filter(substr(CD_SETOR, 1, 2) == "21") %>%
    mutate(pop_setor = coalesce(pop_setor, 0))

  ma_tracts = st_read(MA_MESH, quiet = TRUE) %>%
    select(CD_SETOR, CD_MUN) %>%
    st_make_valid() %>%
    st_transform(proj_wk)

  ma_tracts$tract_area = as.numeric(st_area(ma_tracts))

  # MA municipalities cut by Brazilian Legal Amazon border 
  ma_share = mun_poly %>%
    filter(codmun %/% 100000 == 21) %>%
    mutate(area_clipped_km2 = as.numeric(st_area(.)) / 1e6) %>%
    st_drop_geometry() %>%
    select(codmun, area_clipped_km2) %>%
    left_join(
      ma_tracts %>%
        st_drop_geometry() %>%
        group_by(codmun = as.numeric(CD_MUN)) %>%
        summarise(area_full_km2 = sum(tract_area) / 1e6, .groups = "drop"),
      by = "codmun"
    ) %>%
    mutate(clipped_share = area_clipped_km2 / area_full_km2)
  excess = ma_share$clipped_share[ma_share$clipped_share > 1]
  cut_threshold = 1 - max(quantile(excess, .99, na.rm = TRUE) - 1, 0.02) * 2
  cut_ids = ma_share$codmun[!is.na(ma_share$clipped_share) &
                              ma_share$clipped_share < cut_threshold]

  cut_polys = mun_poly %>%
    filter(codmun %in% cut_ids) %>%
    select(codmun)
  pop_ma = st_intersection(
    ma_tracts[lengths(st_intersects(ma_tracts, cut_polys)) > 0,
              c("CD_SETOR", "tract_area")],
    cut_polys)

  pop_ma$a_in = as.numeric(st_area(pop_ma))
  pop_ma = pop_ma %>%
    st_drop_geometry() %>%
    group_by(codmun, CD_SETOR, tract_area) %>%
    summarise(a_in = sum(a_in), .groups = "drop") %>%
    mutate(ashare = pmin(1, a_in / tract_area)) %>%
    left_join(setor_pop, by = "CD_SETOR") %>%
    mutate(pop_setor = coalesce(pop_setor, 0)) %>%
    group_by(codmun) %>%
    summarise(pop_areawt = sum(pop_setor * ashare), .groups = "drop") %>%
    mutate(pop_areawt = ceiling(pop_areawt))

  saveRDS(pop_ma, partial_population_path)

  rm(setor_pop, ma_tracts, ma_share, excess, cut_threshold, cut_ids, cut_polys,
     pop_ma, csv_in_zip, CENSUS_DIR, MA_MESH, AGG_ZIP, fetch_once)
}

partial_population = readRDS(partial_population_path) %>%
  select(codmun, pop_areawt)


population_legal_amazon = municipal_population %>%
  left_join(partial_population, by = "codmun") %>%
  transmute(codmun, pop_areawt = coalesce(pop_areawt, pop))

rm(partial_population_path, mun_pos, mun_poly, municipal_population, partial_population)


# IBGE 2017 immediate geographic region lookup (microregions)
microreg_lookup = readxl::read_excel(
  paste0(DRAW, "IBGE/regioes_geograficas_composicao_por_municipios_2017_20180911.xlsx")
) %>%
  transmute(
    codmun = as.numeric(CD_GEOCODI),
    microreg = as.character(cod_rgi)
  ) %>%
  distinct(codmun, .keep_all = TRUE)
```


## 1. BUILDING MUNICIPAL PANEL

```{r 1}
# IBGE municipal-seat coordinates; the merged unit uses Santarem's seat.
localidades_zip = paste0(DRAW, "IBGE/Localidades_Municipios_kml.zip")
localidades_files = unzip(localidades_zip, list = TRUE)$Name
localidades_files = localidades_files[
  grepl("_localidades_2022\\.kml$", localidades_files)
]
localidades_codes = as.numeric(sub(
  ".*_(\\d{7})_localidades_2022\\.kml$",
  "\\1",
  localidades_files
))

panel_codes = zones[[which(zones_prefixes == "mun")]]$zone %>%
  st_drop_geometry() %>%
  transmute(codmun = as.numeric(codmun)) %>%
  distinct() %>%
  arrange(codmun) %>%
  pull(codmun)

seat_index = match(panel_codes, localidades_codes)

read_municipal_seat = function(file, codmun) {
  seat = st_read(
    paste0("/vsizip/", localidades_zip, "/", file),
    quiet = TRUE
  ) %>%
    st_zm() %>%
    st_transform(4674)

  coordinates = st_coordinates(seat)[1, ]

  tibble(
    codmun = codmun,
    lat = coordinates[["Y"]],
    lon = coordinates[["X"]]
  )
}

latlong = Map(
  read_municipal_seat,
  localidades_files[seat_index],
  panel_codes
) %>%
  bind_rows() %>%
  arrange(codmun)


rm(localidades_files, localidades_codes, panel_codes, seat_index, read_municipal_seat)


## Distances to Brasilia and nearest coast
# Geodesic distances from each IBGE municipal seat, in kilometers.
sf_use_s2(TRUE)

seat_points = latlong %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4674, remove = FALSE)

# South America coastline (National Water and Sanitation Agency - ANA)
sa_coast = st_read(
  paste0(DRAW, "ANA/geoft_bho_2017_linha_costa.gpkg"),
  layer = "geoft_bho_2017_linha_costa"
) %>%
  st_make_valid() %>%
  st_transform(4674) %>%
  st_make_valid() %>%
  .[1, ] # only continental part of South America

# Brasilia shapefile - capital of Brazil
brasilia = st_read(
  paste0("/vsizip/", localidades_zip, "/kml/DF/brasilia_5300108_localidades_2022.kml"),
  quiet = TRUE) %>%
  st_zm() %>%
  st_transform(4674)

dist = seat_points %>%
  transmute(
    codmun,
    dist_coast = as.numeric(st_distance(geometry, sa_coast) / 1000),
    dist_brasilia = as.numeric(st_distance(geometry, brasilia) / 1000)
  ) %>%
  st_drop_geometry()


rm(sa_coast, brasilia, seat_points, localidades_zip)


## Transportation Infrastructure ##
# PAV, DUP, and EOD are DNIT's paved-road surface codes
ROAD_PAVED = c("PAV", "DUP", "EOD")

# Duplicates Removal
ROAD_TOL_M = 50       # two lines within 50 meters are the same roadbed
ROAD_MIN_RUN = 3      # if the are other three consecutive near-points
ROAD_SPACING_M = 100  # spaced by 100 meters

road_read = function(zip_path, layer, surf_col) {
  st_read(paste0("/vsizip/", zip_path, "/", layer), quiet = TRUE) %>%
    rename(surf = all_of(surf_col)) %>%
    filter(as.character(surf) %in% ROAD_PAVED) %>%
    st_make_valid() %>%
    st_transform(crs = proj_wk) %>%
    st_make_valid()
}

# Clip to municipalities and measure each piece, dropping within-source
# duplicates first so a duplicated roadbed is not clipped twice.
road_clip = function(x, mun_layer) {
  duplicate = duplicated(vapply(
    sf::st_as_binary(sf::st_geometry(x)),
    function(b) paste0(as.character(b), collapse = ""),
    character(1)
  ))
  x[!duplicate, ] %>%
    st_intersection(mun_layer) %>%
    st_make_valid() %>%
    mutate(len_m = as.numeric(st_length(geometry))) %>%
    filter(len_m > 0)
}

road_mun_layer = zones[[which(zones_prefixes == "mun")]]$zone %>% select(codmun)

fed_roads = road_read(paste0(DRAW, "DNIT/202201B.zip"), "SNV_202201B.shp",
                      "ds_sup_fed")
sta_roads = road_read(paste0(DRAW, "DNIT/vw_cide_rod_2021.zip"),
                      "vw_cide_rod_2021.shp", "Superficie")

fed_clip = road_clip(fed_roads, road_mun_layer)
sta_clip = road_clip(sta_roads, road_mun_layer)
rm(fed_roads, sta_roads)
gc()

# Buffering the federal lines lets st_intersects use a spatial index.
sta_lines = st_cast(st_geometry(sta_clip), "LINESTRING", warn = FALSE)
n_pts = pmax(1L, as.integer(round(as.numeric(st_length(sta_lines)) /
                                    ROAD_SPACING_M)))
sample_pts = st_line_sample(sta_lines, n = n_pts)
sample_sf = st_sf(
  piece = rep(seq_len(nrow(sta_clip)), lengths(sample_pts)),
  geometry = st_cast(sample_pts, "POINT")
)
near_fed = lengths(st_intersects(
  sample_sf, st_buffer(st_geometry(fed_clip), ROAD_TOL_M)
)) > 0

by_piece = split(near_fed, sample_sf$piece)
frac = vapply(by_piece, function(h) {
  if (!any(h)) return(0)
  if (length(h) < ROAD_MIN_RUN) return(if (all(h)) 1 else 0)
  runs = rle(h)
  sum(runs$lengths[runs$values & runs$lengths >= ROAD_MIN_RUN]) / length(h)
}, numeric(1))[as.character(seq_len(nrow(sta_clip)))]
frac[is.na(frac)] = 0
sta_clip$len_keep = sta_clip$len_m * (1 - as.numeric(frac))

rm(sta_lines, n_pts, sample_pts, sample_sf, near_fed, by_piece, frac)
gc()

# Federal kilometres in full, state kilometres net of the shared stretches.
roadway = bind_rows(
  st_drop_geometry(fed_clip) %>% transmute(codmun, km = len_m / 1000),
  st_drop_geometry(sta_clip) %>% transmute(codmun, km = len_keep / 1000)
) %>%
  group_by(codmun) %>%
  summarise(way_road_pav = sum(km), .groups = "drop")

roadway = zones[[which(zones_prefixes == "mun")]]$zone %>%
  select(codmun) %>%
  st_drop_geometry() %>%
  left_join(roadway, by = "codmun") %>%
  mutate(way_road_pav = replace_na(way_road_pav, 0))

rm(fed_clip, sta_clip, road_mun_layer)
gc()


# Intersect operating railways with municipalities once, then measure the split
# pieces to avoid carrying pre-split lengths into the municipal totals.
mun_layer = zones[[which(zones_prefixes == "mun")]]$zone %>% select(codmun)

# Transliteration avoids source-encoding drift in the operating-status label.
railway = st_read(
  paste0("/vsizip/", DRAW, "DNIT/BaseFerro.zip/BaseFerro/BaseFerro.shp"),
  quiet = TRUE
) %>%
  filter(grepl("opera", tolower(iconv(as.character(tip_situac),
                                      to = "ASCII//TRANSLIT")))) %>%
  st_make_valid() %>%
  st_transform(crs = proj_wk) %>%
  st_make_valid() %>%
  st_intersection(mun_layer) %>%
  st_make_valid() %>%
  mutate(km = as.numeric(units::set_units(st_length(geometry), "km"))) %>%
  st_drop_geometry() %>%
  group_by(codmun) %>%
  summarise(way_rail = sum(km), .groups = "drop")

rm(mun_layer)


# 2021 municipal mobile coverage
mobile = read_csv2(unz(paste0(DRAW, "ANATEL/cobertura_movel.zip"), 
                       "Atributos_Setores_Censo_2010.csv")) %>%
  select(
    setor = `Código Setor Censitário`,
    area = `Área (km2)`,
    codmun = `Código Município`
  ) %>%
  left_join(
    read_csv2(unz(paste0(DRAW, "ANATEL/cobertura_movel.zip"), 
                  "Cobertura_2021_11_Setores.csv")) %>%
      filter(Operadora == "Todas") %>%
      select(setor = `Código Setor Censitário`, coverarea = Cobertura_Todas),
    by = "setor"
  ) %>%
  arrange(setor) %>%
  mutate(
    setor = as.character(setor),
    coverarea = ifelse(is.na(coverarea), 0, coverarea)
  ) %>%
  group_by(codmun) %>%
  summarise(coverarea = weighted.mean(x = coverarea, w = area))


# Polygon used by both connectivity maps.
legal_amazon_ufs = c("ac", "am", "ap", "ma", "mt", "pa", "ro", "rr", "to")
mobile_map_boundary = mun_zone %>%
  summarise(geometry = st_union(geometry)) %>%
  st_make_valid()
mobile_coverage_2021 = lapply(legal_amazon_ufs, function(uf_i) {
  st_read(paste0("/vsizip/", DRAW, "ANATEL/areas_cobertas.zip/",
                 "todas_todas_", uf_i, "_municipio_simple.kml"),
    quiet = TRUE
  ) %>%
    select(geometry) %>%
    mutate(uf = toupper(uf_i))
}) %>%
  bind_rows() %>%
  st_transform(st_crs(mobile_map_boundary)) %>%
  st_make_valid() %>%
  st_filter(mobile_map_boundary, .predicate = st_intersects) %>%
  select(uf) %>%
  st_transform(4674)

saveRDS(mobile_coverage_2021, paste0(DV, "mobile_coverage_2021.RDS"), 
        compress = "xz")
rm(legal_amazon_ufs, mobile_map_boundary, mobile_coverage_2021)


# Tenure areas
uc_area = zones[[which(zones_prefixes == "uc")]]$zone %>%
  group_by(codmun) %>%
  summarise(
    areauc = sum(areauc),
    geometry = st_union(geometry)
  ) %>%
  ungroup() %>%
  st_drop_geometry()

ti_area = zones[[which(zones_prefixes == "ti")]]$zone %>%
  group_by(codmun) %>%
  summarise(
    areati = sum(areati),
    geometry = st_union(geometry)
  ) %>%
  ungroup() %>%
  st_drop_geometry()

sett_area = zones[[which(zones_prefixes == "sett")]]$zone %>%
  group_by(codmun) %>%
  summarise(
    areasett = sum(areasett),
    geometry = st_union(geometry)
  ) %>%
  ungroup() %>%
  st_drop_geometry()

ndpf_area = zones[[which(zones_prefixes == "ndpf")]]$zone %>%
  group_by(codmun) %>%
  summarise(
    areandpf = sum(areandpf),
    geometry = st_union(geometry)
  ) %>%
  ungroup() %>%
  st_drop_geometry()


# Joining datasets
mun = zones[[which(zones_prefixes == "mun")]]$zone %>%
  st_drop_geometry() %>%
  select(-namemun) %>%
  merge(2017:2025) %>%
  rename(year = y) %>%
  relocate(year, .after = codmun) %>%
  left_join(microreg_lookup, by = "codmun") %>%
  left_join(mobile, by = "codmun") %>%
  left_join(dist, by = "codmun") %>%
  left_join(latlong, by = "codmun") %>%
  left_join(roadway, by = "codmun") %>%
  left_join(railway, by = "codmun") %>%
  mutate(across(c(way_road_pav, way_rail), ~ replace_na(.x, 0))) %>%
  left_join(population_legal_amazon, by = "codmun") %>%
  # Population density uses the population and area inside the panel boundary.
  mutate(dens = pop_areawt / areamun) %>%
  select(-pop_areawt) %>%
  relocate(dens, .after = pop) %>%
  left_join(ti_area, by = "codmun") %>%
  left_join(uc_area, by = "codmun") %>%
  left_join(sett_area, by = "codmun") %>%
  left_join(ndpf_area, by = "codmun") %>%
  left_join(sicar_area, by = "codmun") %>%
  mutate(across(starts_with("area"), ~ replace_na(.x, 0))) %>%
  arrange(codmun, year)

saveRDS(mun, paste0(DV, "munic.RDS"))

rm(mun, mobile, dist, latlong, microreg_lookup, population_legal_amazon, ti_area, uc_area,
   sett_area, ndpf_area, sicar_area, roadway, railway)
```


## 2. FOREST DEGRADATION

```{r 2}
# DETER forest degradation alerts
deter = st_read(paste0(paste0("/vsizip/", DRAW), DETER_zip, "/deter-amz-deter-public.shp")) %>%
  rename(class = CLASSNAME) %>%
  select(-GEOCODIBGE) %>%
  mutate(
    year = year(VIEW_DATE),
    class = case_when(
      class == "CICATRIZ_DE_QUEIMADA" ~ "deter_degr_fire",
      class == "DEGRADACAO" ~ "deter_degr_outr",
      class == "CS_DESORDENADO" ~ "deter_degr_wood",
      class == "CS_GEOMETRICO" ~ "deter_degr_wood",
      class == "CORTE_SELETIVO" ~ "deter_degr_wood"
    )
  ) %>%
  filter(year >= 2016 & year <= 2025 & !is.na(class)) %>%
  st_make_valid() %>%
  st_transform(crs = proj_wk) %>%
  st_make_valid() %>%
  mutate(alert_id = row_number())

# Reusable map cache for plotting maps in starlink_results.Rmd
deter_map = deter %>%
  filter(
    (year %in% c(2022L, 2024L)) |
      (year %in% 2021:2024 & class == "deter_degr_fire")
  ) %>%
  lwgeom::st_snap_to_grid(0.01) %>%
  st_make_valid() %>%
  st_intersection(mun_zone) %>%
  keep_polygons() %>%
  mutate(
    year = as.integer(year),
    alert_class = case_when(
      class == "deter_degr_fire" ~ "fire scar",
      class == "deter_degr_wood" ~ "selective logging",
      class == "deter_degr_outr" ~ "other degradation"
    )
  ) %>%
  select(codmun, year, alert_class) %>%
  group_by(codmun, year, alert_class) %>%
  summarise(.groups = "drop") %>%
  st_make_valid() %>%
  keep_polygons() %>%
  st_transform(4674)

saveRDS(deter_map, paste0(DV, "deter_map.RDS"), compress = "xz")
rm(deter_map)


process_zone = function(deter, zone, prefix) {
  area_col = paste0("area_", prefix, "_deter_degr")
  n_col = paste0("n_", prefix, "_deter_degr")

  x = deter %>%
    st_intersection(zone) %>%
    st_make_valid()

  x = x[as.numeric(st_area(st_geometry(x))) > 0, ]

  out = x %>%
    mutate(
      class = ifelse(class == "", NA, class),
      month = month(VIEW_DATE)
    ) %>%
    group_by(codmun, year, month, class) %>%
    summarise(n = n_distinct(alert_id), .groups = "drop") %>%
    mutate(area = as.numeric(units::set_units(st_area(.), km^2))) %>%
    pivot_wider(
      names_from = class,
      values_from = c(area, n),
      names_glue = paste0("{.value}_", prefix, "_{class}")
    ) %>%
    mutate(
      across(matches("^(area_|n_)"), ~ replace_na(.x, 0)),
      !!area_col := rowSums(
        across(starts_with(paste0("area_", prefix, "_deter_degr_"))),
        na.rm = TRUE
      ),
      !!n_col := rowSums(
        across(starts_with(paste0("n_", prefix, "_deter_degr_"))),
        na.rm = TRUE
      )
    ) %>%
    st_drop_geometry() %>%
    group_by(codmun, year, month) %>%
    summarise(across(matches("^(area_|n_)"), sum), .groups = "drop")

  if (prefix != "mun") {
    out = out %>% select(codmun, year, month, all_of(c(area_col, n_col)))
  }
  out
}

active_deter_prefixes = c(
  "mun", "ti", "uc", "sett", "ndpf",
  "sicar21_large", "sicar21_small",
  "over_ndpf_large", "over_ndpf_small"
)
active_deter_zones = zones[zones_prefixes %in% active_deter_prefixes]

options(future.globals.maxSize = 2 * 1024^3)
plan(multisession, workers = min(max_workers, length(active_deter_zones)))

results = future_lapply(
  active_deter_zones,
  function(z) process_zone(deter, z$zone, z$prefix)
)

plan(sequential) # Parallelization (END)

# Recombine municipality batches that share a prefix.
results_combined = list()

active_deter_zone_prefixes = vapply(active_deter_zones, function(z) z$prefix, character(1))
for (p in unique(active_deter_zone_prefixes)) {
  idx = which(active_deter_zone_prefixes == p)
  results_combined[[p]] = bind_rows(results[idx]) %>%
    group_by(codmun, year, month) %>%
    summarise(across(everything(), sum), .groups = "drop")
}

# Now join each element has unique column names
deter = zones[[which(zones_prefixes == "mun")]]$zone %>%
  select(codmun) %>%
  st_drop_geometry() %>%
  merge(2016:2025) %>%
  rename(year = y) %>%
  merge(1:12) %>%
  rename(month = y) %>%
  arrange(codmun, year, month)

for (res in results_combined) {
  deter = deter %>% left_join(res, by = c("codmun", "year", "month"))
}
rm(results)

deter = deter %>%
  mutate_at(
    vars(starts_with("area_"), starts_with("n_")),
    ~ replace(., is.na(.), 0)
  )

saveRDS(
  deter %>%
    filter(year > 2016L | (year == 2016L & month >= 8L), year <= 2021L) %>%
    group_by(codmun) %>%
    summarise(
      pre2022_fire_scar_area = sum(area_mun_deter_degr_fire, na.rm = TRUE),
      .groups = "drop"
    ),
  paste0(DV, "deter_pre2022_fire_scar.RDS")
)

saveRDS(
  deter %>%
    filter(year >= 2017 & year <= 2025) %>%
    group_by(codmun, year) %>%
    summarise(
      across(starts_with("area_") | starts_with("n_"), ~ sum(.x, na.rm = TRUE))
    ) %>%
    ungroup(),
  paste0(DV, "deter_normalyr.RDS")
)

rm(deter, process_zone, results_combined, res,
   active_deter_prefixes, active_deter_zones, active_deter_zone_prefixes)
```


## 3. ENVIRONMENTAL ENFORCEMENT

```{r 3}
## 3.1. IBAMA
ibama = data.frame()

for (t in 2016:2025) {
  x = read.csv2(unz(paste0(DRAW, "IBAMA/auto_infracao_csv.zip"),
                    paste0("auto_infracao_ano_", t, ".csv"))) %>%
    mutate(
      year = year(DAT_HORA_AUTO_INFRACAO),
      month = month(DAT_HORA_AUTO_INFRACAO)
    ) %>%
    select(year, month, INFRACAO_AREA, TIPO_INFRACAO, codmun = COD_MUNICIPIO) %>%
    mutate(
      TIPO_INFRACAO = case_when(
        TIPO_INFRACAO %in% c("Fauna", "Pesca") ~ "Fauna",
        TIPO_INFRACAO %in% c("Flora") ~ "Flora",
        TIPO_INFRACAO %in%
          c("Qualidade Ambiental", "Cadastro Técnico Federal", "Controle Ambiental",
            "Licenciamento") ~ "Pollution",
        TIPO_INFRACAO %in%
          c("Administração Ambiental", "Ord. Urbano e Patr. Cultural",
            "Unidade de Conservação",
            "Org. Gen. Modific. e Biopirataria", "Outras") ~ "Administrative"
      ),
      INFRACAO_AREA = replace_na(INFRACAO_AREA, ""),
      TIPO_INFRACAO = replace_na(TIPO_INFRACAO, ""),
      dummy_fire = ifelse(
        INFRACAO_AREA == "Desmatamento e Queimada" | INFRACAO_AREA == "Queimada",
        1, 0),
      dummy_flor = ifelse(TIPO_INFRACAO == "Flora", 1, 0),
      dummy_faun = ifelse(TIPO_INFRACAO == "Fauna", 1, 0),
      dummy_polu = ifelse(TIPO_INFRACAO == "Pollution", 1, 0),
      dummy_admi = ifelse(TIPO_INFRACAO == "Administrative", 1, 0)
    )

  ibama = bind_rows(
    ibama,
    x %>% select(codmun, year, month, starts_with("dummy_"))
  )
}

ibama = ibama %>%
  group_by(codmun, year, month) %>%
  summarise(
    n_mun_ibama = n(),
    n_mun_ibama_fire = sum(dummy_fire),
    n_mun_ibama_flor = sum(dummy_flor),
    n_mun_ibama_faun = sum(dummy_faun),
    n_mun_ibama_polu = sum(dummy_polu),
    n_mun_ibama_admi = sum(dummy_admi),
    .groups = "drop"
  )


## 3.2. ICMBio ##

# For conservation-unit-specific or unclassified notices, keywords identify the
# three substantive infraction classes. Residuals are classified as Administrative.
rx = list(
  Fauna = c(
    "\\bpesc", "\\bquelonio", "\\bcac(?:a|ar|ou|ando|ador(?:es)?)\\b", "\\barmas? de fogo\\b",
    "\\btatus?\\b", "fuzarca", "zangaria", "espingarda", "\\bredes?\\b", "goiamum", "tartaruga",
    "pirarucu", "peixe-boi", "municao", "alevino", "molinete", "calibre", "armadilha", "abater anima",
    "soltura", "\\bsuino", "animais domesticos", "criacao de peixes"
  ),
  Pollution = c(
    "vazamento", "residuo", "esgoto", "efluente", "derram", "resto", "lixo", "dejeto", "sedimento",
    "afluente", "descarte", "embalagem", "entulho", "toxic", "agrotoxico", "exploracao de ouro",
    "garimpo", "extracao de areia", "retirada de areia", "minerio", "minera", "quartz", "quartiz",
    "garimpagem", "granito", "minerios", "exploracao aurifera", "garimpe", "minwracao", "mercurio",
    "cascalho", "amolar"
  ),
  Flora = c(
    "retroescadeira", "suprimi", "destruir", "ocupacao", "incendio", "vegetacao nativa", "supress",
    "aterr", "estrada", "acesso", "abertura de estrada", "queima", "grilagem", "hectar", "destruicao",
    "eros", "invasao", "\\b(?:ha|\\d+\\s*ha)\\b", "uso de fogo", "marge(m|ns)", "assoreamento",
    "atraves do fogo", "uso do fogo", "abertura de trilha", "escavacao", "limpeza da area",
    "limpeza de area", "loteamento", "trator escavadeira", "lesao", "destruindo", "utilizando fogo",
    "reabertura", "remocao", "abrir uma trilha", "atear fogo", "plato", "retro escavadeira",
    "parcelamento", "desmat", "corte raso", "ramal", "produtos florestais", "motosserra", "corte",
    "madeira", "carread", "derrubada", "exploracao florestal", "angelim", "tora", "carreador",
    "retirada", "palmito", "cortar", "serraria", "trepadeira", "trator esteira", "paraju", "aderno",
    "miridiba", "tauari", "maracatiara", "cedro", "essencia", "madeireiro", "cipo silvestre",
    "madeireira", "araucaria", "abatimento", "transporte de arvore", "seletiva", "castanheira",
    "\\bretirada\\s+de\\s+\\d+(?:[\\.,]\\d+)?\\s*arvore(s)?\\b", "produto florestal",
    "trator de esteira", "trator valmit", "moto-serra", "motoserra", "produtos florestrais",
    "trator de pneu", "toreiro", "trator fiatallis", "\\babater\\s+\\d+\\s+arvore\\b", "pinhao",
    "plantas", "bromelia", "orquid(e|i)a", "orquidia", "acai", "frut", "lenha", "agricultura",
    "agricola", "gado", "pecuaria", "pasto", "bovino", "pastagem"
  )
)
rx = map_chr(rx, ~ str_c(.x, collapse = "|"))

icmbio = st_read(
  paste0("/vsizip/", DRAW, "ICMBio/autos_infracao_icmbio_shp.zip/autos_infracao_icmbio.shp")
) %>%
  st_make_valid() %>%
  st_transform(crs = proj_wk) %>%
  st_make_valid() %>%
  rename(year = ano) %>%
  mutate(year = as.numeric(year)) %>%
  filter(year >= 2016 & year <= 2025) %>%
  mutate(
    month = month(data),
    across(c(tipo_infra, desc_ai), ~ .x %>%
      stri_replace_all_regex("[\\x00-\\x08\\x0B\\x0C\\x0E-\\x1F]", "") %>%
      stri_trans_nfc() %>%
      str_to_lower(locale = "pt") %>%
      stri_trans_general("Latin-ASCII") %>%
      str_squish() %>%
      replace_na("")),
    TIPO_INFRACAO = case_when(
      str_detect(tipo_infra, "contra a flora") ~ "Flora",
      str_detect(tipo_infra, "contra a fauna") ~ "Fauna",
      str_detect(tipo_infra, "poluicao") ~ "Pollution",
      str_detect(tipo_infra, "administrativas|ordenamento urbano") ~ "Administrative",
      str_detect(desc_ai, rx[["Fauna"]]) ~ "Fauna",
      str_detect(desc_ai, rx[["Pollution"]]) ~ "Pollution",
      str_detect(desc_ai, rx[["Flora"]]) ~ "Flora",
      TRUE ~ "Administrative"
    ),

    dummy_fire = as.integer(
      str_detect(desc_ai, "fogo|incendi|queima") &
        !str_detect(
          desc_ai,
          "\\b(?:armas?|espingardas?|armadilhas?)\\s+de\\s+fogo\\b|\\bfogos\\b"
        ) |
        artigo_1 %in% c("58", "58-A", "58-B", "58A", "58B") |
        artigo_2 %in% c("58", "58-A", "58-B", "58A", "58B")
    ),

    dummy_flor = 1 * (TIPO_INFRACAO == "Flora"),
    dummy_faun = 1 * (TIPO_INFRACAO == "Fauna"),
    dummy_polu = 1 * (TIPO_INFRACAO == "Pollution"),
    dummy_admi = 1 * (TIPO_INFRACAO == "Administrative")
  ) %>%
  select(year, month, dummy_fire, dummy_flor,
         dummy_faun, dummy_polu, dummy_admi)

icmbio = icmbio %>%
  st_intersection(
    zones[[which(zones_prefixes == "mun")]]$zone %>% select(codmun)
  ) %>%
  st_drop_geometry() %>%
  group_by(codmun, year, month) %>%
  summarise(
    n_mun_icmbio = n(),
    n_mun_icmbio_fire = sum(dummy_fire),
    n_mun_icmbio_flor = sum(dummy_flor),
    n_mun_icmbio_faun = sum(dummy_faun),
    n_mun_icmbio_polu = sum(dummy_polu),
    n_mun_icmbio_admi = sum(dummy_admi),
    .groups = "drop"
  )


police = full_join(ibama, icmbio, by = c("codmun", "year", "month")) %>%
  mutate(across(starts_with("n_mun_"), ~ replace_na(.x, 0))) %>%
  transmute(
    codmun, year, month,
    n_mun_police = n_mun_ibama + n_mun_icmbio,
    n_mun_police_fire = n_mun_ibama_fire + n_mun_icmbio_fire,
    n_mun_police_flor = n_mun_ibama_flor + n_mun_icmbio_flor,
    n_mun_police_faun = n_mun_ibama_faun + n_mun_icmbio_faun,
    n_mun_police_polu = n_mun_ibama_polu + n_mun_icmbio_polu,
    n_mun_police_admi = n_mun_ibama_admi + n_mun_icmbio_admi
  ) %>%
  group_by(codmun, year) %>%
  summarise(across(starts_with("n_mun_police"), sum), .groups = "drop")

police = zones[[which(zones_prefixes == "mun")]]$zone %>%
  st_drop_geometry() %>%
  select(codmun) %>%
  merge(2017:2025) %>%
  rename(year = y) %>%
  left_join(police, by = c("codmun", "year")) %>%
  mutate(across(starts_with("n_mun_police"), ~ replace_na(.x, 0))) %>%
  arrange(codmun, year)

saveRDS(police, paste0(DV, "police_normalyr.RDS"))

rm(police, ibama, icmbio, x, rx)
```


## 4. SATELLITE BROADBAND

```{r 4}
## 4.1. STARLINK
aux = data.frame()
for (t in 2022:2025) {
  aux = bind_rows(
    aux,
    read_csv2(unz(
      paste0(DRAW, "ANATEL/acessos_banda_larga_fixa.zip"),
      paste0("Acessos_Banda_Larga_Fixa_", t, ".csv")
    )) %>%
      filter(
        Tecnologia == "VSAT" &
          `Meio de Acesso` == "Satélite" &
          Empresa == "STARLINK BRAZIL SERVICOS DE INTERNET LTDA."
      ) %>%
      select(codmun = `Código IBGE Município`, year = Ano, month = Mês, Acessos) %>%

      # Mojui's accesses fold into the merged Santarem unit BEFORE aggregation
      mutate(codmun = merge_codmun(codmun)) %>%
      group_by(codmun, year, month) %>%
      summarise(starlink = sum(Acessos))
  ) %>%
    arrange(codmun, year, month)
}

starlink = zones[[1]]$zone %>%
  st_drop_geometry() %>%
  select(codmun, pop) %>%
  unique() %>%
  merge(2016:2025) %>%
  rename(year = y) %>%
  merge(1:12) %>%
  rename(month = y) %>%
  arrange(codmun, year, month) %>%
  left_join(aux, by = c("codmun", "year", "month")) %>%
  mutate(
    starlink = ifelse(is.na(starlink), 0, starlink),
    starlink = starlink / (pop / 1000)
  ) %>%
  select(-pop)

saveRDS(
  starlink %>%
    filter(year >= 2022 & year <= 2025) %>%
    group_by(codmun, year) %>%
    summarise(starlink = mean(starlink)),
  paste0(DV, "starlink_normalyr.RDS")
)

saveRDS(
  starlink %>%
    mutate(year = ifelse(month >= 8, year + 1, year)) %>%
    filter(year >= 2022 & year <= 2025) %>%
    group_by(codmun, year) %>%
    summarise(starlink = mean(starlink)),
  paste0(DV, "starlink_prodesyr.RDS")
)

## 4.2. ALL OTHER GEO SATELLITE INTERNET PROVIDERS ##
aux = data.frame()

for (t in c("2015-2016", "2017-2018", "2019-2020", "2021", "2022", "2023", "2024", "2025")) {
  aux = bind_rows(
    aux,
    read_csv2(unz(
      paste0(DRAW, "ANATEL/acessos_banda_larga_fixa.zip"),
      paste0("Acessos_Banda_Larga_Fixa_", t, ".csv")
    )) %>%
      filter(
        Tecnologia == "VSAT" &
          `Meio de Acesso` == "Satélite" &
          Empresa != "STARLINK BRAZIL SERVICOS DE INTERNET LTDA."
      ) %>%
      select(codmun = `Código IBGE Município`, year = Ano, month = Mês, Acessos) %>%
      # Mojui's accesses fold into the merged Santarem unit BEFORE aggregation
      mutate(codmun = merge_codmun(codmun)) %>%
      group_by(codmun, year, month) %>%
      summarise(geosat = sum(Acessos))
  ) %>%
    arrange(codmun, year, month)
}

geosat = zones[[1]]$zone %>%
  st_drop_geometry() %>%
  select(codmun, pop) %>%
  unique() %>%
  merge(2016:2025) %>%
  rename(year = y) %>%
  merge(1:12) %>%
  rename(month = y) %>%
  arrange(codmun, year, month) %>%
  left_join(aux, by = c("codmun", "year", "month")) %>%
  mutate(
    geosat = ifelse(is.na(geosat), 0, geosat),
    geosat = geosat / (pop / 1000)
  ) %>%
  select(-pop)

saveRDS(
  geosat %>%
    filter(year >= 2017 & year <= 2025) %>%
    group_by(codmun, year) %>%
    summarise(geosat = mean(geosat)),
  paste0(DV, "geosat_normalyr.RDS")
)

saveRDS(
  geosat %>%
    mutate(year = ifelse(month >= 8, year + 1, year)) %>%
    filter(year >= 2017 & year <= 2025) %>%
    group_by(codmun, year) %>%
    summarise(geosat = mean(geosat)),
  paste0(DV, "geosat_prodesyr.RDS")
)

rm(aux, geosat, starlink)
```

## 5. AIR POLLUTION

```{r 5}
# (a) Build task list: split each PM size into n_chunks temporal chunks
n_chunks = 3

tasks = list()
for (PM_SIZE in c("1", "2.5", "10")) {
  N = sub("\\.", "p", PM_SIZE)
  PM_clean = sub("\\.", "", PM_SIZE)

  # Read layer metadata only (dates)
  aux = rast(paste0(DRAW, "CAMS/data_sfc.nc"), subds = paste0("pm", N))
  dates = time(aux)

  # Resume check: find which months still need processing.
  files = list.files(DV, pattern = paste0("pm", PM_clean, "mun"))

  month_id = format(dates, "%Y-%m")
  completed_months = if (length(files) > 0) {
    basename(files) %>%
      sub(paste0("pm", PM_clean, "mun_"), "", x = ., fixed = TRUE) %>%
      sub(".RDS", "", x = ., fixed = TRUE) %>%
      sub("_", "-", x = ., fixed = TRUE)
  } else {
    character(0)
  }
  keep = which(!month_id %in% completed_months)

  if (length(keep) > 0) {
    # Keep every calendar month inside one task so parallel workers never write
    # partial versions of the same monthly cache.
    layers_by_month = split(keep, month_id[keep])
    n_groups = min(n_chunks, length(layers_by_month))
    month_chunks = if (n_groups == 1) {
      list(seq_along(layers_by_month))
    } else {
      split(
        seq_along(layers_by_month),
        cut(seq_along(layers_by_month), n_groups, labels = FALSE)
      )
    }
    chunks = lapply(
      month_chunks,
      function(i) unname(unlist(layers_by_month[i], use.names = FALSE))
    )

    for (ch in seq_along(chunks)) {
      tasks[[length(tasks) + 1]] = list(
        PM_SIZE = PM_SIZE, N = N, PM_clean = PM_clean, layers = chunks[[ch]]
      )
    }
  } else {
    message(paste0("PM", PM_SIZE, " already complete - skipping"))
  }
}

rm(aux)


# (b) Run extraction in parallel - one task per chunk
amz_nogeo = zones[[1]]$zone %>% st_drop_geometry() %>% select(codmun)

{
  # Parallelization
  options(future.globals.maxSize = 4 * 1024^3)
  plan(multisession, workers = max(1, min(max_workers, length(tasks))))

  future_lapply(
    tasks,
    function(task) {
      library(terra)
      library(lubridate)
      library(exactextractr)
      library(dplyr)
      library(sf)
      library(tibble)

      PM_SIZE = task$PM_SIZE
      N = task$N
      PM_clean = task$PM_clean
      lyrs = task$layers

      # Each worker reloads the raster (terra is not serializable)
      r = rast(paste0(DRAW, "CAMS/data_sfc.nc"), subds = paste0("pm", N))

      all_dates = time(r)

      dates = all_dates[lyrs]
      r = r[[lyrs]] %>%
        project(proj_wk) %>%
        crop(zones[[1]]$zone, snap = "out")

      month_id = format(dates, "%Y-%m")
      month_levels = unique(month_id)
      r = tapp(r, index = match(month_id, month_levels), fun = mean, na.rm = TRUE)
      dates = as.Date(paste0(month_levels, "-01"))

      for (idx in seq_len(nlyr(r))) {
        tif = r[[idx]]
        y = year(dates[idx])
        m = month(dates[idx])

        ext = exact_extract(
          tif, zones[[1]]$zone, coverage_area = TRUE, progress = FALSE,
          max_cells_in_memory = 3e+9, summarize_df = TRUE, fun = sum_coverage
        )

        bind_cols(
          amz_nogeo, year = y, month = m, pm_mun = sapply(ext, sum, na.rm = TRUE)
        ) %>%
          saveRDS(paste0(DV, "pm", PM_clean, "mun_", y, "_", sprintf("%02d", m), ".RDS"))
      }

      invisible(NULL)
    },
    future.seed = TRUE
  )

  plan(sequential)
}


# (c) Joining all RDS into final dataset
pm_files = list.files(DV, pattern = "^pm\\d+mun_", full.names = TRUE)

pm_mun = map_dfr(pm_files, function(f) {
  # Extract PM size from filename (e.g., "pm25mun_2017_01.RDS" â†’ "2.5")
  pm_code = sub("^pm(\\d+)mun_.*", "\\1", basename(f))
  readRDS(f) %>%
    mutate(
      pm_mun = pm_mun * 1e9, # kg/m3 -> ug/m3
      pmsize = pm_code
    )
})

pm_mun = pm_mun %>%
  group_by(codmun, year, month, pmsize) %>%
  summarise(pm_mun = mean(pm_mun), .groups = "drop") %>%
  pivot_longer(cols = starts_with("pm_m")) %>%
  mutate(name = paste0("pm", pmsize, substr(name, 4, nchar(name)))) %>%
  select(-pmsize) %>%
  pivot_wider()

saveRDS(
  pm_mun %>%
    filter(year >= 2017 & year <= 2025) %>%
    group_by(codmun, year) %>%
    summarise(
      pm1mun = mean(pm1mun),
      pm10mun = mean(pm10mun),
      pm25mun = mean(pm25mun)
    ) %>%
    ungroup(),
  paste0(DV, "pmcams_normalyr.RDS")
)

rm(list = intersect(
  c("pm_mun", "pm_files", "amz_nogeo", "tasks", "n_chunks", "chunks"),
  ls()
))
```

## 6. MAX TEMPERATURE AND PRECIPITATION

```{r 6}
bbox = st_bbox(zones[[1]]$zone %>% st_transform(4326)) + c(-.5, -.5, .5, .5)
amz_nogeo = zones[[1]]$zone %>% st_drop_geometry() %>% select(codmun)

# Generic worker function - each file is one task
extract_raster_file = function(
  filepath, varname, yr_start, yr_len, mo_start, mo_len, nodata = NULL
) {
  library(terra)
  library(exactextractr)
  library(sf)
  library(dplyr)

  tif = rast(filepath) %>% crop(bbox)
  if (!is.null(nodata)) {
    tif[tif == nodata] = NA
  }
  tif = tif %>% project(proj_wk)

  f = basename(filepath)
  y = as.numeric(substr(f, yr_start, yr_start + yr_len - 1))
  m = as.numeric(substr(f, mo_start, mo_start + mo_len - 1))

  ext = exact_extract(
    tif, zones[[1]]$zone, coverage_area = TRUE, progress = FALSE,
    summarize_df = TRUE, fun = sum_coverage
  )

  bind_cols(amz_nogeo, year = y, month = m, !!varname := sapply(ext, sum, na.rm = TRUE))
}


## 6.1. Maximum temperature (CHIRTS-ERA5) ##
keep_panel_climate_months = function(paths) {
  file_ym = str_match(basename(paths), "([0-9]{4})\\.([0-9]{2})\\.tiff?$")
  file_month = as.Date(paste(file_ym[, 2], file_ym[, 3], "01", sep = "-"))
  paths[
    !is.na(file_month) &
      file_month >= as.Date("2016-08-01") &
      file_month <= as.Date("2025-07-31")
  ]
}

tmax_files = list.files(paste0(DRAW, "CHC/CHIRTS-ERA5_Tmax"), pattern = "\\.tif$",
                        full.names = TRUE) %>% keep_panel_climate_months()

{
  # Parallelization
  options(future.globals.maxSize = 2 * 1024^3)
  plan(multisession, workers = max(1, min(max_workers, length(tmax_files))))

  tmax_list = future_lapply(tmax_files, future.seed = TRUE, function(fp) {
    extract_raster_file(
      fp, "tmax", yr_start = 26, yr_len = 4,
      mo_start = 31, mo_len = 2, nodata = -9999
    )
  })

  plan(sequential)
}

tmax = bind_rows(tmax_list) %>% arrange(codmun, year, month)

saveRDS(
  tmax %>%
    filter(year >= 2017 & year <= 2024) %>%
    group_by(codmun, year) %>%
    summarise(tmax = mean(tmax)) %>%
    ungroup() %>%
    arrange(codmun, year),
  paste0(DV, "tmax_normalyr.RDS")
)

saveRDS(
  tmax %>%
    mutate(year = ifelse(month >= 8, year + 1, year)) %>%
    filter(year >= 2017 & year <= 2025) %>%
    group_by(codmun, year) %>%
    summarise(tmax = mean(tmax)) %>%
    ungroup() %>%
    arrange(codmun, year),
  paste0(DV, "tmax_prodesyr.RDS")
)

rm(tmax, tmax_list)


## 6.2. Precipitation (CHIRPS v3.0) ##
prcp_files = list.files(paste0(DRAW, "CHC/CHIRPS-v3_latam"), pattern = "\\.tiff?$",
                        full.names = TRUE) %>% keep_panel_climate_months()

plan(multisession, workers = max(1, min(max_workers, length(prcp_files))))

prcp_list = future_lapply(prcp_files, future.seed = TRUE, function(fp) {
  extract_raster_file(
    fp, "prcp", yr_start = 13, yr_len = 4,
    mo_start = 18, mo_len = 2, nodata = -9999
  )
})

plan(sequential)

prcp = bind_rows(prcp_list) %>% arrange(codmun, year, month)

saveRDS(
  prcp %>%
    filter(year >= 2017 & year <= 2024) %>%
    group_by(codmun, year) %>%
    summarise(prcp = mean(prcp)) %>%
    ungroup() %>%
    arrange(codmun, year),
  paste0(DV, "prcp_normalyr.RDS")
)

saveRDS(
  prcp %>%
    mutate(year = ifelse(month >= 8, year + 1, year)) %>%
    filter(year >= 2017 & year <= 2025) %>%
    group_by(codmun, year) %>%
    summarise(prcp = mean(prcp)) %>%
    ungroup() %>%
    arrange(codmun, year),
  paste0(DV, "prcp_prodesyr.RDS")
)

rm(prcp, prcp_list, bbox, amz_nogeo, sum_coverage, extract_raster_file,
   keep_panel_climate_months)
```

## 7. SPEI-12 (Copernicus ERA5-Drought)

```{r 7}
# Copernicus ERA5-Drought SPEI-12
spei_zip = paste0(DRAW, "CAMS/SPEI-12_Amazon_2017_2025.zip")

# The archive holds one NetCDF per month, named ..._ref1991to2020_YYYYMM....nc
spei_members = tibble(member = unzip(spei_zip, list = TRUE)$Name) %>%
  filter(grepl("^SPEI12_", member)) %>%
  mutate(
    year_month = sub(".*_ref1991to2020_([0-9]{6}).*", "\\1", member),
    year = as.integer(substr(year_month, 1, 4)),
    month = as.integer(substr(year_month, 5, 6))
  )

spei_tasks = bind_rows(
  spei_members %>%
    filter(month == 12, year >= 2017, year <= 2024) %>%
    mutate(anchor = "dec"),
  spei_members %>%
    filter(month == 7, year >= 2017, year <= 2025) %>%
    mutate(anchor = "jul")
) %>%
  arrange(anchor, year)

# Unzip the 17 endpoint files to a temporary folder instead.
spei_dir = tempfile("spei12_")
dir.create(spei_dir)
unzip(spei_zip, files = spei_tasks$member, exdir = spei_dir)
spei_tasks$path = file.path(spei_dir, spei_tasks$member)

spei_mun = readRDS(paste0(DV, "mun_zone.RDS")) %>%
  select(codmun) %>%
  arrange(codmun)
spei_crs = st_crs(spei_mun)$wkt
spei_bbox = st_bbox(st_transform(spei_mun, 4326)) + c(-0.5, -0.5, 0.5, 0.5)

# Coverage-area-weighted mean over the cells with a valid SPEI value
spei_weighted_mean = function(values, coverage_area) {
  valid = is.finite(values) & is.finite(coverage_area) & coverage_area > 0
  if (!any(valid)) {
    return(NA_real_)
  }
  sum(values[valid] * coverage_area[valid]) / sum(coverage_area[valid])
}


extract_spei_file = function(i) {
  library(terra)
  library(exactextractr)
  library(sf)
  library(dplyr)

  spei = rast(spei_tasks$path[i]) %>%
    crop(ext(spei_bbox))
  # -9999 is the only fill value. Finite extreme values are part of the
  # provider's fitted index and are kept.
  spei[spei <= -9990] = NA
  spei = project(spei, spei_crs)

  tibble(
    codmun = as.numeric(spei_mun$codmun),
    year = spei_tasks$year[i],
    anchor = spei_tasks$anchor[i],
    spei = as.numeric(exact_extract(
      spei, spei_mun, spei_weighted_mean, coverage_area = TRUE, progress = FALSE
    ))
  )
}


options(future.globals.maxSize = 2 * 1024^3)
plan(multisession, workers = min(max_workers, nrow(spei_tasks)))
spei_anchors = future_lapply(seq_len(nrow(spei_tasks)), extract_spei_file, future.seed = TRUE) %>%
  bind_rows()
plan(sequential)
unlink(spei_dir, recursive = TRUE)


spei12_normalyr = spei_anchors %>%
  filter(anchor == "dec") %>%
  transmute(codmun, year, spei12_dec = spei) %>%
  arrange(codmun, year)
spei12_prodesyr = spei_anchors %>%
  filter(anchor == "jul") %>%
  transmute(codmun, year, spei12_jul = spei) %>%
  arrange(codmun, year)

saveRDS(spei12_normalyr, paste0(DV, "spei12_copernicus_era5_normalyr.RDS"))
saveRDS(spei12_prodesyr, paste0(DV, "spei12_copernicus_era5_prodesyr.RDS"))

rm(
  spei_zip, spei_members, spei_tasks, spei_dir, spei_mun, spei_crs, spei_bbox,
  spei_weighted_mean, extract_spei_file, spei_anchors, spei12_normalyr,
  spei12_prodesyr
)
```

## 8. FOREST COVER AND TRANSITIONS

```{r 8}
# Rasters are in EPSG 4326, so the extraction polygons must be too.
mun = zones[[which(zones_prefixes == "mun")]]$zone %>%
  st_make_valid() %>% st_transform(4326) %>% st_make_valid()

# Land cover class groupings
forest_classes = c(1, 3, 4, 5, 6, 49)
natural_nonforest = c(10, 11, 12, 32, 29, 50, 13)
agriculture_pec = c(15)
agriculture_agr = c(18, 19, 39, 20, 40, 62, 41, 36, 46, 47, 35, 48)
agriculture_out = c(14, 9, 21)
nonveg_min = c(30)
nonveg_urb = c(24)
other_destination = c(22, 23, 25, 75, 26, 33, 31)

make_transitions = function(from, to, label) {
  expand.grid(from = from, to = to) %>%
    mutate(code = from * 100 + to, category = label)
}

transition_table = bind_rows(
  make_transitions(forest_classes, forest_classes, "sh_flo"),
  make_transitions(forest_classes, natural_nonforest, "sh_veg"),
  make_transitions(forest_classes, agriculture_pec, "sh_agr_pec"),
  make_transitions(forest_classes, agriculture_agr, "sh_agr_agr"),
  make_transitions(forest_classes, agriculture_out, "sh_agr_out"),
  make_transitions(forest_classes, nonveg_min, "sh_nvg_min"),
  make_transitions(forest_classes, nonveg_urb, "sh_nvg_urb"),
  make_transitions(forest_classes, other_destination, "sh_other")
) %>%
  distinct(code, category) %>%
  mutate(cat_id = as.integer(factor(category)))

# All category labels in cat_id order (for column naming)
cat_labels = transition_table %>%
  distinct(cat_id, category) %>%
  arrange(cat_id) %>%
  pull(category)
output_cat_labels = setdiff(cat_labels, "sh_other")
shm_labels = sub("^sh_", "shm_", output_cat_labels)

n_cats = length(cat_labels)
reclass_mat = as.matrix(transition_table[, c("code", "cat_id")])


# MapBiomas raster file paths
period = 2016:2024
flist = file.path(DRAW, "MapBiomas", paste0("brazil_coverage_", period, ".tif"))
period_actual = period


# -----------------------------------------------------------------------
# Transition function: returns forest-origin transition shares with two bases:
# sh_*  = transition cells / prior-year forest-origin cells
# shm_* = transition cells / total valid municipal cells
# -----------------------------------------------------------------------
raster_trans = function(raster_ini, raster_end, poly) {
  transition_raster = raster_ini * 100 + raster_end
  transition_category_raster = classify(transition_raster, rcl = reclass_mat, others = NA)

  ext = exact_extract(
    transition_category_raster, poly, progress = FALSE,
    fun = function(values, coverage_fraction) as.list(table(values)),
    max_cells_in_memory = 3e+10
  )

  total_cells = exact_extract(
    raster_ini, poly, progress = FALSE,
    fun = function(values, coverage_fraction) sum(!is.na(values)),
    max_cells_in_memory = 3e+10
  )

  df = tibble(id = seq_along(ext), values = ext) %>%
    unnest_wider(values, names_sep = "_")

  required_cols = paste0("values_", seq_len(n_cats))
  missing_cols = setdiff(required_cols, names(df))
  if (length(missing_cols) > 0) {
    df[missing_cols] = 0
  }

  df = df %>%
    select(id, all_of(required_cols)) %>%
    mutate(across(
      all_of(required_cols),
      ~ replace_na(as.numeric(unlist(.x)), 0)
    ))

  colnames(df) = c("id", cat_labels)

  df = df %>%
    mutate(
      sh_total = rowSums(across(all_of(cat_labels)), na.rm = TRUE),
      total_cells = as.numeric(total_cells)
    )

  df = df %>%
    mutate(across(
      all_of(output_cat_labels),
      ~ .x / total_cells,
      .names = "municipal_{.col}"
    )) %>%
    rename_with(~ sub("^municipal_sh_", "shm_", .x), starts_with("municipal_sh_")) %>%
    mutate(across(all_of(cat_labels), ~ .x / sh_total)) %>%
    select(id, all_of(output_cat_labels), all_of(shm_labels))

  df = poly %>%
    select(codmun) %>%
    st_drop_geometry() %>%
    bind_cols(df) %>%
    select(-id)

  return(df)
}


# -----------------------------------------------------------------------
# Main loop: parallelized by state
# -----------------------------------------------------------------------
uflist = sort(unique(substr(mun$codmun, 1, 2)))

completed_ufs = uflist[
  file.exists(paste0(DV, "mb_trans_", uflist, "_", period_actual[1], "-",
                     period_actual[length(period_actual)], ".RDS")) &
    file.exists(paste0(DV, "mb_forest_", uflist, ".RDS"))
]
if (length(completed_ufs) > 0) {
  message("MapBiomas states already complete: ", paste(completed_ufs, collapse = ", "))
}
uflist = setdiff(uflist, completed_ufs)

options(future.globals.maxSize = 4 * 1024^3)
{
  plan(multisession, workers = max(1, min(max_workers, 3, length(uflist))))

  invisible(future_lapply(seq_along(uflist), future.seed = TRUE, function(i) {
    library(terra)
    library(exactextractr)
    library(sf)
    library(dplyr)
    library(tidyr)
    library(purrr)

    uf = uflist[i]
    temp_mun = mun[substr(mun$codmun, 1, 2) == uf, ]
    temp_bbox = st_bbox(temp_mun) + c(-.5, -.5, .5, .5)

    n_rasters = length(flist)

    load_raster = function(k) {
      r = rast(flist[k]) %>% crop(temp_bbox)
      mask(r, temp_mun)
    }

    forest_stats = function(r, year_value) {
      forest_count = exactextractr::exact_extract(
        r, temp_mun, function(values, coverage_fraction) sum(values %in% forest_classes),
        progress = FALSE,
        max_cells_in_memory = 3e+09
      )

      total_count = exactextractr::exact_extract(
        r, temp_mun, function(values, coverage_fraction) sum(!is.na(values)),
        progress = FALSE,
        max_cells_in_memory = 3e+09
      )

      tibble(
        codmun = temp_mun$codmun,
        year = year_value,
        area_floresta = as.numeric(forest_count),
        area_total = as.numeric(total_count),
        p_forest = area_floresta / area_total
      )
    }

    trans = data.frame()
    forest_list = vector("list", n_rasters)
    r_prev = load_raster(1)
    forest_list[[1]] = forest_stats(r_prev, period_actual[1])

    for (j in 2:n_rasters) {
      r_curr = load_raster(j)
      forest_list[[j]] = forest_stats(r_curr, period_actual[j])

      aux = raster_trans(raster_ini = r_prev, raster_end = r_curr, poly = temp_mun) %>%
        mutate(year = period_actual[j])
      trans = rbind(trans, aux)

      # Shift: current becomes previous, free memory
      r_prev = r_curr
      rm(r_curr)
      gc()
    }

    rm(r_prev)
    gc()

    saveRDS(
      trans,
      paste0(DV, "mb_trans_", uf, "_", period_actual[1], "-", period_actual[n_rasters], ".RDS")
    )

    forest = bind_rows(forest_list) %>% arrange(codmun, year)

    saveRDS(forest, paste0(DV, "mb_forest_", uf, ".RDS"))

    invisible(NULL)
  }))

  plan(sequential)
}

# -----------------------------------------------------------------------
# Consolidate state-level files
# -----------------------------------------------------------------------
mbtrans = list.files(
  DV, paste0("^mb_trans_[0-9]{2}_", first(period_actual), "-", last(period_actual), "\\.RDS$"),
  full.names = TRUE
) %>%
  map_dfr(readRDS) %>%
  select(codmun, year, all_of(output_cat_labels), all_of(shm_labels)) %>%
  arrange(codmun, year)

saveRDS(mbtrans, paste0(DV, "mbtrans.RDS"))

# Match only the per-state forest-stock files.
mb_forest = list.files(DV, "^mb_forest_[0-9]{2}\\.RDS$", full.names = TRUE) %>%
  map_dfr(readRDS) %>%
  arrange(codmun, year)
saveRDS(mb_forest, paste0(DV, "mb_forest.RDS"))

rm(
  mbtrans, mb_forest, mun, uflist, completed_ufs, flist, period, period_actual,
  transition_table, reclass_mat, cat_labels, output_cat_labels, shm_labels, n_cats,
  forest_classes, natural_nonforest, agriculture_pec, agriculture_agr, agriculture_out,
  nonveg_min, nonveg_urb, other_destination, make_transitions, raster_trans
)
```

## 9. FIRE IGNITIONS

```{r 9}
uf_codes = c(
  "ACRE" = "AC", "AMAPÁ" = "AP", "AMAZONAS" = "AM", "MARANHÃO" = "MA", "MATO GROSSO" = "MT",
  "PARÁ" = "PA", "RONDÔNIA" = "RO", "RORAIMA" = "RR", "TOCANTINS" = "TO"
)

# Begins in 2019, since NOAA-20 has been available from this year onward
hotspots = map_dfr(2019:2025, function(yr) {
  csv_name = paste0(if (yr == 2025) "tmp/" else "", "focos_br_todos-sats_", yr, ".csv")
  read_csv(
    unz(paste0(DRAW, "INPE/focos_br_todos-sats_", yr, ".zip"), csv_name),
    show_col_types = FALSE
  ) %>%
    filter(
      estado %in% names(uf_codes),
      satelite %in% c("NOAA-20", "NPP-375", "NPP-375D")
    ) %>%
    transmute(
      latitude, longitude, data_pas = with_tz(data_pas, tzone = "America/Manaus"),
      satelite, sigla_uf = unname(uf_codes[estado])
    )
})

hotspots = split(hotspots, hotspots$sigla_uf)

# Reuse per-state ignition caches; delete one to rebuild that state.
to_cluster = hotspots[
  !file.exists(file.path(DV, paste0("igni_", names(hotspots), ".RDS")))
]

if (length(to_cluster) == 0) {
  message("ignitions: all ", length(hotspots),
          " per-state files already present -- skipping spotoroo clustering")
} else {
  message("ignitions: clustering ", length(to_cluster), " of ", length(hotspots),
          " states (", paste(names(to_cluster), collapse = ", "), ")")
  options(future.globals.maxSize = 2 * 1024^3)
  plan(multisession, workers = min(max_workers, length(to_cluster)))

  future_lapply(
    to_cluster,
    future.seed = TRUE,
    future.packages = "spotoroo",
    function(uf_hotspots) {
      uf = uf_hotspots$sigla_uf[1]

      ignitions_uf = spotoroo::hotspot_cluster(
        uf_hotspots, lon = "longitude", lat = "latitude", obsTime = "data_pas",
        activeTime = 120, adjDist = 750, minPts = 5, minTime = 3,
        ignitionCenter = "mean", timeUnit = "h", timeStep = 1
      )$ignition

      saveRDS(ignitions_uf, paste0(DV, "igni_", uf, ".RDS"))
      invisible(NULL)
    }
  )

  plan(sequential)
}


# Combine cached ignitions and project them to the municipal CRS.
ignitions = map_dfr(names(hotspots), function(uf) {
  readRDS(paste0(DV, "igni_", uf, ".RDS"))
}) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(st_crs(zones[[1]]$zone))


# Unbiased temporal and 10-km-grid spatial HHIs.
igni_pts = st_join(
  ignitions %>% select(obsTime),
  zones[[1]]$zone %>% select(codmun),
  join = st_within
) %>%
  filter(!is.na(codmun))

conc_igni = igni_pts %>%
  st_drop_geometry() %>%
  transmute(codmun, year = year(obsTime), date = as_date(obsTime)) %>%
  filter(year >= 2017, year <= 2025) %>%
  count(codmun, year, date, name = "n_day") %>%
  group_by(codmun, year) %>%
  summarise(
    n = sum(n_day),
    pairs = sum(n_day * (n_day - 1)),
    .groups = "drop"
  ) %>%
  transmute(
    codmun, year,
    conc_igni_hhiunb = if_else(n > 1, pairs / (n * (n - 1)), NA_real_)
  )

igni_xy = st_coordinates(igni_pts)
disp_igni = igni_pts %>%
  st_drop_geometry() %>%
  transmute(
    codmun,
    year = year(obsTime),
    cell_x = floor(igni_xy[, 1] / 10000),
    cell_y = floor(igni_xy[, 2] / 10000)
  ) %>%
  filter(year >= 2017, year <= 2025) %>%
  count(codmun, year, cell_x, cell_y, name = "n_cell") %>%
  group_by(codmun, year) %>%
  summarise(
    n = sum(n_cell),
    pairs = sum(n_cell * (n_cell - 1)),
    .groups = "drop"
  ) %>%
  transmute(
    codmun, year,
    disp_shhiunb = if_else(n > 1, pairs / (n * (n - 1)), NA_real_)
  )

saveRDS(conc_igni, paste0(DV, "conc_igni_normalyr.RDS"))
saveRDS(disp_igni, paste0(DV, "disp_igni_normalyr.RDS"))

# Count ignitions by municipality and month using the same spatial join.
igni_mun = igni_pts %>%
  st_drop_geometry() %>%
  mutate(year = year(obsTime), month = month(obsTime)) %>%
  group_by(codmun, year, month) %>%
  summarise(n_igni = n()) %>%
  ungroup()

igni_mun = zones[[1]]$zone %>%
  select(codmun) %>%
  st_drop_geometry() %>%
  merge(2019:2025) %>%
  rename(year = y) %>%
  merge(1:12) %>%
  rename(month = y) %>%
  left_join(igni_mun, by = join_by(codmun, year, month)) %>%
  mutate(n_igni = ifelse(is.na(n_igni), 0, n_igni)) %>%
  arrange(codmun, year, month)


# Saving processed data
saveRDS(
  igni_mun %>%
    filter(year >= 2017 & year <= 2025) %>%
    group_by(codmun, year) %>%
    summarise(n_igni = sum(n_igni)),
  paste0(DV, "ignimun_normalyr.RDS")
)

rm(
  uf_codes, hotspots, to_cluster, ignitions, igni_pts, igni_xy,
  conc_igni, disp_igni, igni_mun
)
```

## 10. CRIMINAL FACTIONS

```{r 10}
# Pool the two report editions into one time-invariant, ever-documented presence measure.
factions_long = readr::read_csv(
  paste0(DRAW, "FBSP/amazon_factions_2023_2024.csv"),
  show_col_types = FALSE
) %>%
  filter(report_year %in% c(2023L, 2024L)) %>%
  transmute(
    codmun = as.numeric(codmun),
    faction
  ) %>%
  # Mojui re-keyed to the merged Santarem unit
  mutate(codmun = merge_codmun(codmun)) %>%
  unique()


# Collapse to a cross-section. The resulting faction variables do not vary by panel year
factions = factions_long %>%
  distinct(codmun) %>%
  mutate(faction = 1L)


# Build the complete municipal skeleton locally so this chunk is self-contained.
factions = zones[[which(zones_prefixes == "mun")]]$zone %>%
  st_drop_geometry() %>%
  select(codmun) %>%
  left_join(factions, by = "codmun") %>%
  mutate(faction = ifelse(is.na(faction), 0, faction))

saveRDS(factions, paste0(DV, "factions.RDS"))
rm(factions_long, factions)
```

## 11. COMMODITY PRICES

```{r 11-commodity-prices}
# SEAB-PR / SIMA daily quotations
SIMA_ARCHIVE_PATHS = setNames(
  file.path(paste0(DRAW, "SEAB-PR"),
            c("sima_2016.rar", "sima_2017.rar", paste0("sima_", 2018:2024, ".zip"))),
  2016:2024
)

KEEP = "^Boi|^Soja"

# one sheet -> tidy rows. Identical to the original: skip 3, rename the first and
# antepenultimate columns, take the value one row down, and keep cattle and soy.
sima_sheet = function(file, sheet, dat) {
  aux = suppressMessages(read_excel(file, sheet = sheet, skip = 3))
  if (ncol(aux) < 3) return(NULL)

  # Trim trailing all-zero or all-blank filler before locating the "Media"
  junk = vapply(aux, function(c) {
    v = suppressWarnings(as.numeric(as.character(c)))
    all(is.na(c) | trimws(as.character(c)) == "") || all(is.na(v) | v == 0)
  }, logical(1))

  # Removing "junk" XLS sheets
  keep = which(!junk)
  if (!length(keep)) return(NULL)
  aux = aux[, seq_len(max(keep)), drop = FALSE]
  if (ncol(aux) < 3) return(NULL)
  colnames(aux)[c(1, ncol(aux) - 2)] = c("commodity", "price")
  aux %>%
    mutate(price = as.numeric(lead(price, 1))) %>%
    filter(grepl(KEEP, commodity)) %>%
    select(commodity, price) %>%
    mutate(commodity = word(commodity, 1),
           year = year(dat), month = month(dat), day = day(dat))
}

sima_files = function(y) {
  archive_path = SIMA_ARCHIVE_PATHS[[as.character(y)]]
  extract_dir = tempfile(pattern = paste0("seab-pr-", y, "-"))
  dir.create(extract_dir)
  archive::archive_extract(archive_path, dir = extract_dir)

  files = list.files(extract_dir, pattern = "\\.xls[xm]?$", recursive = TRUE, full.names = TRUE)
  # A few archives contain print or copy duplicates that would otherwise
  # double-weight the corresponding business day.
  files = files[!grepl("impressao|c[oó]pia", files, ignore.case = TRUE)]
  if (length(files) == 0) {
    stop("No Excel workbooks found inside ", basename(archive_path))
  }
  list(files = files, extract_dir = extract_dir)
}

# Daily files are named DD-MM-YYYY from April 2023 onward but DD-MM-YY in January to March 2023.
sima_file_date = function(b) {
  if (grepl("^[0-9]{2}-[0-9]{2}-[0-9]{4}", b))
    return(as.Date(substr(b, 1, 10), format = "%d-%m-%Y"))
  if (grepl("^[0-9]{2}-[0-9]{2}-[0-9]{2}([^0-9]|$)", b))
    return(as.Date(substr(b, 1, 8), format = "%d-%m-%y"))
  as.Date(NA)
}

cmdty_prices = tibble()
failed = character()

for (ano in 2016:2024) {
  sima_year = sima_files(ano)
  files = sima_year$files
  message(sprintf("SIMA %d: %d files", ano, length(files)))
  for (f in files) {
    b = basename(f)
    fdate = sima_file_date(b)
    got = try({
      if (!is.na(fdate)) {
        sima_sheet(f, 1, fdate)                       # one day per file
      } else {                                        # multi-sheet workbook
        bind_rows(lapply(excel_sheets(f), function(s) {

          # February 2019 was silently lost. Parse per sheet, inside try().
          d = try({
            hdr = colnames(suppressMessages(read_excel(f, sheet = s)))
            as.Date(hdr[length(hdr) - 1], tryFormats = c("%d-%m-%Y", "%d/%m/%Y"))
          }, silent = TRUE)
          if (inherits(d, "try-error") || length(d) == 0 || is.na(d)) return(NULL)
          # sima_sheet() is wrapped too: one unreadable sheet must cost a day only
          r = try(sima_sheet(f, s, d), silent = TRUE)
          if (inherits(r, "try-error")) NULL else r
        }))
      }
    }, silent = TRUE)
    if (inherits(got, "try-error") || is.null(got)) failed = c(failed, f)
    else cmdty_prices = bind_rows(cmdty_prices, got)
  }
  unlink(sima_year$extract_dir, recursive = TRUE, force = TRUE)
}
message(sprintf("parsed %d daily quotations; %d files failed", nrow(cmdty_prices), length(failed)))
if (length(failed)) message("  first failures: ", paste(head(basename(failed), 5), collapse = ", "))

# Safety net. Removing double-count.
cmdty_prices = cmdty_prices %>% distinct(commodity, year, month, day, .keep_all = TRUE)

cmdty_monthly = cmdty_prices %>%
  mutate(year = ifelse(year < 100, year + 2000, year)) %>%
  group_by(commodity, year, month) %>%
  summarise(price = mean(price, na.rm = TRUE), n_days = n(), .groups = "drop")

# Deflate with monthly IPCA variation from IBGE SIDRA table 1737, variable 63.
ipca_path = paste0(DRAW, "IBGE/ipca_mensal_sidra.csv")
if (!file.exists(ipca_path)) {
  message("fetching IPCA from SIDRA table 1737 ...")
  dir.create(dirname(ipca_path), recursive = TRUE, showWarnings = FALSE)
  sidra_url = "https://apisidra.ibge.gov.br/values/t/1737/n1/all/v/63/p/201601-202412"
  sidra_resp = httr::GET(sidra_url, httr::timeout(180))
  sidra_ipca = jsonlite::fromJSON(
    httr::content(sidra_resp, "text", encoding = "UTF-8"),
    simplifyDataFrame = TRUE
  )[-1, ]
  ipca_out = tibble(
    period = sidra_ipca$D3C,
    variacao_mensal = suppressWarnings(as.numeric(sidra_ipca$V))
  ) %>%
    filter(!is.na(variacao_mensal)) %>%
    mutate(
      ano = as.integer(substr(period, 1, 4)),
      mes = as.integer(substr(period, 5, 6))
    ) %>%
    arrange(ano, mes)
  ipca_out = ipca_out %>% select(ano, mes, variacao_mensal)
  write.csv(ipca_out, ipca_path, row.names = FALSE)
  rm(sidra_url, sidra_resp, sidra_ipca, ipca_out)
}

ipca = read_csv(ipca_path, show_col_types = FALSE) %>%
  filter(!is.na(variacao_mensal)) %>%
  arrange(ano, mes) %>%
  mutate(index = cumprod(1 + variacao_mensal / 100)) %>%
  filter(ano >= 2016, ano <= 2024) %>%
  transmute(year = as.numeric(ano), month = as.numeric(mes),
            factor = last(index) / index)          # to December 2024 prices

cmdty = cmdty_monthly %>%
  left_join(ipca, by = c("year", "month")) %>%
  mutate(price = price * factor,
         commodity = recode(commodity, Boi = "cattle", Soja = "soy")) %>%
  select(-factor)

# Annual calendar-year prices.
cmdty_annual = cmdty %>%
  filter(year >= 2016, year <= 2024) %>%
  group_by(commodity, year) %>% summarise(price = mean(price), .groups = "drop") %>%
  pivot_wider(names_from = "commodity", names_prefix = "price_", values_from = "price")

cmdty_price_panel = full_join(
  cmdty_annual %>%
    transmute(
      year = as.integer(year),
      soy_price_current = price_soy,
      cattle_price_current = price_cattle
    ),
  cmdty_annual %>%
    transmute(
      year = as.integer(year) + 1L,
      soy_price_lagged_level = price_soy,
      cattle_price_lagged_level = price_cattle
    ),
  by = "year"
) %>%
  filter(year >= 2017L, year <= 2025L) %>%
  arrange(year)

saveRDS(cmdty_price_panel, paste0(DV, "cmdty_price_panel.RDS"))

rm(
  cmdty_monthly, ipca, cmdty_annual, cmdty_price_panel,
  SIMA_ARCHIVE_PATHS, sima_year
)
```



## 12. POTENTIAL SOY YIELD

```{r 12-potential-soy-yield}
# GAEZ v5 (historical 2001-2020)
amz = zones[[which(zones_prefixes == "mun")]]$zone %>%
  select(codmun) %>%
  st_make_valid()

# Use only the full-input, high-input/rain-fed soy potential raster.
soy = exactextractr::exact_extract(
  terra::rast(file.path(
    DRAW, "FAO",
    "DATA_GAEZ-V5_MAPSET_RES05-YXX_GAEZ-V5.RES05-YXX.HP0120.AGERA5.HIST.SOY.HRLM.tif"
  )) %>% terra::project(proj_wk),
  amz, coverage_area = TRUE, summarize_df = TRUE,
  fun = function(x) {
    list(x %>%
      mutate(
        coverage_area = coverage_area * 1e-4,
        total_area = sum(coverage_area, na.rm = TRUE),
        yield = (value / 1000) * (coverage_area / total_area)
      ) %>%
      select(yield))
  }
)

soy = tibble(
  codmun = amz$codmun,
  soyield_ful_mun_high = round(
    vapply(soy, function(x) sum(x[["yield"]], na.rm = TRUE), numeric(1)),
    4
  )
)

saveRDS(soy, paste0(DV, "soy_potential.RDS"))
rm(amz, soy)
```

## 13. PASTURE SUITABILITY

```{r 13-pasture-suitability}
# FGGD global-land pasture suitability, FAO catalog
pasture_zip = normalizePath(file.path(DRAW, "FAO", "Map6_56.zip"), winslash = "/", mustWork = TRUE)
past = terra::rast(paste0("/vsizip/", pasture_zip, "/pasture_si/hdr.adf"))

# -997 is a nodata sentinel, not a suitability value (1.2% of Amazon-area cells)
past = terra::classify(past, cbind(-997, NA))

# Native values from 0 to 10000 encode the official continuous 0-100 index
past = terra::project(past, proj_wk)

# Area-weighted municipal mean
amz = zones[[which(zones_prefixes == "mun")]]$zone %>%
  select(codmun) %>%
  st_make_valid()

pasture = tibble(
  codmun = amz$codmun,
  pasture_suit_mun = round(
    exactextractr::exact_extract(
      past, amz, fun = "weighted_mean", weights = "area", progress = FALSE
    ),
    4
  )
)

saveRDS(pasture, paste0(DV, "pasture_suitability.RDS"))
rm(amz, past, pasture, pasture_zip)
```

## APPENDIX: MORTALITY RATES

```{r appendix-mortality}
mun = zones[[1]]$zone %>%
  select(codmun, pop) %>%
  mutate(codmun6 = as.numeric(substr(codmun, 1, 6))) %>%
  st_drop_geometry()

deaths = tibble()
files = c(
  "Mortalidade_Geral_2017_csv.zip", "Mortalidade_Geral_2018_csv.zip",
  "Mortalidade_Geral_2019_csv.zip", "Mortalidade_Geral_2020_csv.zip",
  "Mortalidade_Geral_2021_csv.zip", "DO22OPEN_csv.zip", "DO23OPEN_csv.zip", "DO24OPEN_csv.zip"
)

for (file in files) {
  if (substr(file, nchar(file) - 2, nchar(file)) == "zip") {
    aux = read_csv2(unz(
      paste0(paste0(DRAW, "DATASUS/"), file),
      gsub("_(?=[^_]*$)", ".", gsub("\\.zip$", "", file), perl = TRUE)
    ))
  } else {
    aux = read_csv2(paste0(paste0(DRAW, "DATASUS/"), file))
  }

  aux = aux %>%
    mutate(
      year = as.numeric(substr(DTOBITO, 5, 8)),
      month = as.numeric(substr(DTOBITO, 3, 4)),
      cid = toupper(CAUSABAS),
      cid_l = substr(cid, 1, 1),
      cid_n = as.numeric(substr(cid, 2, 3)),
      type = case_when(
        # http://tabnet.datasus.gov.br/cgi/sih/mxcid10lm.htm
        cid_l == "A" | cid_l == "B" ~ "infect",
        cid_l == "C" | (cid_l == "D" & cid_n <= 48) ~ "neoplasm",
        cid_l == "D" & cid_n >= 49 ~ "blood",
        cid_l == "E" ~ "endocrine",
        cid_l == "F" ~ "mental",
        cid_l == "G" ~ "nervous",
        cid_l == "H" & cid_n <= 59 ~ "eyes",
        cid_l == "H" & cid_n >= 60 ~ "ears",
        cid_l == "I" ~ "circulatory",
        cid_l == "J" | (cid_l == "U" & cid_n == 4) ~ "respiratory",
        cid_l == "K" ~ "digestive",
        cid_l == "L" ~ "skin",
        cid_l == "M" ~ "musculoskeletal",
        cid_l == "N" ~ "genitourinary",
        cid_l == "O" ~ "pregnancy",
        cid_l == "P" ~ "perinatal",
        cid_l == "Q" ~ "malformation",
        cid_l == "R" ~ "other",
        cid_l %in% c("S", "T", "V", "W", "X", "Y") ~ "extcause",
        cid_l == "Z" | cid_l == "U" ~ "other"
      )
    ) %>%
    rename(codmun6 = CODMUNOCOR) %>%
    # Joining Santarém + Mojuí dos Campos
    mutate(codmun6 = ifelse(as.numeric(codmun6) == MOJUI_CODMUN %/% 10,
                            SANTAREM_CODMUN %/% 10, as.numeric(codmun6))) %>%
    left_join(mun, by = "codmun6")

  deaths = bind_rows(
    deaths,
    # Total Deaths
    aux %>%
      mutate(type = "total") %>%
      group_by(codmun, year, month, type) %>%
      summarise(qty = n()) %>%
      ungroup(),
    # Deaths by ICD
    aux %>%
      group_by(codmun, year, month, type) %>%
      summarise(qty = n()) %>%
      ungroup()
  )
}

deaths = mun %>%
  merge(2017:2025) %>% rename(year = y) %>%
  merge(1:12) %>% rename(month = y) %>%
  left_join(deaths, by = c("codmun", "year", "month")) %>%
  mutate(qty = qty / (pop / 1e5)) %>%
  filter(!is.na(codmun) & !is.na(type)) %>%
  pivot_wider(
    names_from = "type",
    names_prefix = "mort_",
    values_from = "qty",
    values_fill = 0
  ) %>%
  select(codmun, year, month, mort_total, starts_with("mort_"))

saveRDS(
  deaths %>%
    filter(year >= 2017 & year <= 2025) %>%
    group_by(codmun, year) %>%
    summarise(
      across(starts_with("mort_"), ~ sum(.x, na.rm = TRUE))
    ) %>%
    ungroup(),
  paste0(DV, "deaths_normalyr.RDS")
)

rm(deaths, aux, mun)
```

## APPENDIX: PRODES DEFORESTATION

```{r appendix-prodes}
# PRODES deforestation 6.25 ha or higher, Legal Amazon product.
prodes = st_read(
  paste0("/vsizip/", DRAW, "INPE/yearly_deforestation_amazonia_legal.zip/",
         "yearly_deforestation_amazonia_legal.shp")
) %>%
  rename(class = sub_class) %>%
  mutate(
    class = case_when(
      class == "corte raso com solo exposto" ~ "prodes_defo_cces",
      class == "corte raso com vegetação" ~ "prodes_defo_ccwv",
      class == "mineração" ~ "prodes_defo_mine",
      class == "desmatamento por degradação progressiva" ~ "prodes_defo_dpdp",
      TRUE ~ "prodes_defo_NA"
    )
  ) %>%
  filter(year >= 2017 & year <= 2025) %>%
  st_make_valid() %>%
  st_transform(crs = proj_wk) %>%
  st_make_valid() %>%
  select(year, class) %>%
  mutate(poly_id = row_number())

prodes = prodes %>%
  st_intersection(zones[[which(zones_prefixes == "mun")]]$zone %>% select(codmun)) %>%
  st_make_valid()
prodes = prodes[as.numeric(st_area(st_geometry(prodes))) > 0, ] %>%
  group_by(codmun, year, class) %>%
  summarise(n = n_distinct(poly_id), .groups = "drop") %>%
  mutate(area = as.numeric(units::set_units(st_area(.), km^2))) %>%
  pivot_wider(
    names_from = class, values_from = c(area, n),
    names_glue = "{.value}_mun_{class}"
  ) %>%
  mutate(
    across(matches("^(area_|n_)"), ~ replace_na(.x, 0)),
    area_mun_prodes_defo = rowSums(across(starts_with("area_mun_prodes_defo_")), na.rm = TRUE),
    n_mun_prodes_defo = rowSums(across(starts_with("n_mun_prodes_defo_")), na.rm = TRUE)
  ) %>%
  st_drop_geometry() %>%
  group_by(codmun, year) %>%
  summarise(across(matches("^(area_|n_)"), sum), .groups = "drop")

prodes = zones[[which(zones_prefixes == "mun")]]$zone %>%
  select(codmun) %>%
  st_drop_geometry() %>%
  merge(2017:2025) %>%
  rename(year = y) %>%
  left_join(prodes, by = c("codmun", "year")) %>%
  mutate(across(matches("^(area_|n_)"), ~ replace_na(.x, 0))) %>%
  arrange(codmun, year)

saveRDS(prodes, paste0(DV, "prodes_prodesyr.RDS"))

rm(prodes)
```

## 14. JOINING DATASETS

```{r 14-joining-datasets}
# Normal year
ds_normalyr = readRDS(paste0(DV, "munic.RDS")) %>%
  left_join(
    readRDS(paste0(DV, "mb_forest.RDS")) %>%
      filter(year == 2021) %>%
      select(codmun, p_forest_2021 = p_forest),
    by = "codmun"
  ) %>%
  left_join(
    readRDS(paste0(DV, "mb_forest.RDS")) %>%
      transmute(codmun, year = year + 1L, p_forest_lag = p_forest),
    by = c("codmun", "year")
  ) %>%
  mutate(
    area_forest = areamun * p_forest_2021,
    lag_areaforest = areamun * p_forest_lag
  ) %>%
  select(-p_forest_2021, -p_forest_lag) %>%
  relocate(area_forest, lag_areaforest, .after = areamun) %>%
  left_join(readRDS(paste0(DV, "starlink_normalyr.RDS")), by = c("codmun", "year")) %>%
  mutate(reg_sample = ifelse(year > 2021, 1, 0), .after = year) %>%
  left_join(readRDS(paste0(DV, "geosat_normalyr.RDS")), by = c("codmun", "year")) %>%
  left_join(readRDS(paste0(DV, "soy_potential.RDS")), by = "codmun") %>%
  left_join(readRDS(paste0(DV, "pasture_suitability.RDS")), by = "codmun") %>%
  left_join(readRDS(paste0(DV, "cmdty_price_panel.RDS")), by = "year") %>%
  left_join(readRDS(paste0(DV, "tmax_normalyr.RDS")), by = c("codmun", "year")) %>%
  left_join(readRDS(paste0(DV, "prcp_normalyr.RDS")), by = c("codmun", "year")) %>%
  left_join(
    readRDS(file.path(DV, "spei12_copernicus_era5_normalyr.RDS")),
    by = c("codmun", "year")
  ) %>%
  left_join(readRDS(paste0(DV, "deter_pre2022_fire_scar.RDS")), by = "codmun") %>%
  mutate(pre2022_fire_scar_intensity = 100 * pre2022_fire_scar_area / area_forest,
         pre2022_fire_scar_area = NULL) %>%
  left_join(readRDS(paste0(DV, "factions.RDS")), by = "codmun") %>%
  left_join(readRDS(paste0(DV, "deter_normalyr.RDS")), by = c("codmun", "year")) %>%
  left_join(readRDS(paste0(DV, "mbtrans.RDS")), by = c("codmun", "year")) %>%
  left_join(readRDS(paste0(DV, "police_normalyr.RDS")), by = c("codmun", "year")) %>%
  left_join(readRDS(paste0(DV, "pmcams_normalyr.RDS")), by = c("codmun", "year")) %>%
  left_join(readRDS(paste0(DV, "deaths_normalyr.RDS")), by = c("codmun", "year")) %>%
  left_join(readRDS(paste0(DV, "ignimun_normalyr.RDS")), by = c("codmun", "year")) %>%
  left_join(readRDS(paste0(DV, "conc_igni_normalyr.RDS")), by = c("codmun", "year")) %>%
  left_join(readRDS(paste0(DV, "disp_igni_normalyr.RDS")), by = c("codmun", "year")) %>%
  filter(year <= 2024) %>%
  arrange(codmun, year)

saveRDS(ds_normalyr, paste0(DATA_DIR, "dataset_normalyr.RDS"))


# PRODES year
ds_prodesyr = readRDS(paste0(DV, "munic.RDS")) %>%
  left_join(
    readRDS(paste0(DV, "mb_forest.RDS")) %>%
      filter(year == 2021) %>%
      select(codmun, p_forest_2021 = p_forest),
    by = "codmun"
  ) %>%
  mutate(area_forest = areamun * p_forest_2021) %>%
  select(
    -pop, -p_forest_2021,
    -matches("^area(ti|uc|sett|ndpf|sicar|over_)")
  ) %>%
  relocate(area_forest, .after = areamun) %>%
  left_join(readRDS(paste0(DV, "starlink_prodesyr.RDS")), by = c("codmun", "year")) %>%
  mutate(reg_sample = ifelse(year > 2022, 1, 0), .after = year) %>%
  left_join(readRDS(paste0(DV, "geosat_prodesyr.RDS")), by = c("codmun", "year")) %>%
  left_join(readRDS(paste0(DV, "soy_potential.RDS")), by = "codmun") %>%
  left_join(readRDS(paste0(DV, "pasture_suitability.RDS")), by = "codmun") %>%
  left_join(
    readRDS(paste0(DV, "cmdty_price_panel.RDS")) %>%
      select(year, soy_price_lagged_level, cattle_price_lagged_level),
    by = "year"
  ) %>%
  left_join(readRDS(paste0(DV, "tmax_prodesyr.RDS")), by = c("codmun", "year")) %>%
  left_join(readRDS(paste0(DV, "prcp_prodesyr.RDS")), by = c("codmun", "year")) %>%
  left_join(
    readRDS(file.path(DV, "spei12_copernicus_era5_prodesyr.RDS")),
    by = c("codmun", "year")
  ) %>%
  left_join(readRDS(paste0(DV, "deter_pre2022_fire_scar.RDS")), by = "codmun") %>%
  mutate(pre2022_fire_scar_intensity = 100 * pre2022_fire_scar_area / area_forest,
         pre2022_fire_scar_area = NULL) %>%
  left_join(
    readRDS(paste0(DV, "prodes_prodesyr.RDS")) %>%
      select(codmun, year, matches("^(area|n)_mun_prodes_defo"), -ends_with("_NA")),
    by = c("codmun", "year")
  ) %>%
  mutate(
    area_mun_prodes_defo_ccut = coalesce(area_mun_prodes_defo_cces, 0) +
      coalesce(area_mun_prodes_defo_ccwv, 0),
    n_mun_prodes_defo_ccut = coalesce(n_mun_prodes_defo_cces, 0) + coalesce(n_mun_prodes_defo_ccwv, 0)
  ) %>%
  filter(year <= 2025) %>%
  arrange(codmun, year)

saveRDS(ds_prodesyr, paste0(DATA_DIR, "dataset_prodesyr.RDS"))
```

## Session information

```{r SessionInfo}
sessionInfo()
```
